---
title: "Lab 13 — Partial pooling"
subtitle: "Compute and visualize shrinkage on synthetic section data, then check it against the closed form"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 13 — Hierarchical models](../notes/week-13-hierarchical-models.qmd). There you wrote the
> two-level model and the partial-pooling formula
> $\hat\theta_j^{\text{PP}} = \lambda_j\,\bar y_j + (1-\lambda_j)\,\mu$. Here you *produce* the three
> estimates — no pooling, complete pooling, partial pooling — from seeded synthetic section data,
> visualize the shrinkage, and verify your simulated partial-pooled mean matches the closed-form
> value.

## Goal

Compare no-pooling, complete-pooling, and partial-pooling estimates on synthetic section data and
visualize shrinkage. By the end you will have (1) the three estimates per section in a table, (2) a
shrinkage plot connecting each section's no-pooling mean to its partial-pooled mean, and (3) a
numeric check that the closed-form partial-pooled mean agrees with a simulated posterior mean — all
reproduced from a seeded, re-runnable `.qmd`.

## Setup

First, make sure your local toolchain is ready: R, VS Code, and Quarto, following the
[R + VS Code + Quarto setup page](../resources/r-vscode-quarto-setup.qmd). Then, create a new file
named `lab-13.qmd`, paste the chunks below in order, and render top to bottom. Every chunk uses
**base R only** — no add-on packages — so it runs on a clean install. Seeds are fixed so your
numbers match the text.

## Steps

### Step 1 — Simulate synthetic section data

We invent five course sections of mixed size with known within-section noise `sigma` and a known
between-section spread `tau`. Keeping the truth known lets you later check that partial pooling
behaves the way the formula predicts. These are the same settings as the Week 13 note.

```{r}
#| label: lab13-simulate
set.seed(1313)
sigma <- 10        # within-section sd (known)
tau   <- 6         # between-section sd (known)
sizes <- c(8, 30, 12, 6, 20)
labels <- c("A", "B", "C", "D", "E")
true_means <- c(70, 78, 73, 84, 76)   # synthetic section truths

# store each section's raw scores in a list
scores <- vector("list", length(sizes))
for (j in seq_along(sizes)) scores[[j]] <- rnorm(sizes[j], true_means[j], sigma)
sapply(scores, length)                 # confirm the section sizes
```

### Step 2 — The three estimates

Compute the no-pooling mean (each section's own average), the complete-pooling mean (one overall
average), and the partial-pooled mean (the precision-weighted blend from Week 13).

```{r}
#| label: lab13-estimates
ybar <- sapply(scores, mean)                       # no pooling
grand <- sum(sizes * ybar) / sum(sizes)            # complete pooling (overall mean)
lambda <- (sizes / sigma^2) / (sizes / sigma^2 + 1 / tau^2)
pp <- lambda * ybar + (1 - lambda) * grand         # partial pooling

data.frame(section = labels, n = sizes,
           no_pooling = round(ybar, 2),
           complete_pooling = round(grand, 2),
           lambda = round(lambda, 2),
           partial_pooled = round(pp, 2))
```

Read the table: every `partial_pooled` value sits between its `no_pooling` value and the
`complete_pooling` overall mean, and `lambda` (the weight on the section's own data) is smallest for
the smallest sections.

### Step 3 — Visualize the shrinkage

Draw the shrinkage plot: each section's no-pooling mean on the left, its partial-pooled mean on the
right, joined by a segment, with the overall mean as a dashed reference. Steeper segments are
sections that shrank more.

```{r}
#| label: fig-lab13-shrinkage
#| fig-cap: "Shrinkage plot for five synthetic sections: no-pooling means (left) connected to partial-pooled means (right), with the overall mean dashed. Small sections shrink more than large ones."
#| fig-alt: "Two columns of points, no pooling on the left and partial pooling on the right, with a dashed horizontal line near 75 for the overall mean. Segments connect each section's two estimates; segments for small sections slope steeply toward the dashed line while the large section's segment is nearly flat."
plot(NA, xlim = c(0.7, 2.3), ylim = range(c(ybar, pp, grand)),
     xaxt = "n", xlab = "", ylab = "estimated section mean score",
     main = "Lab 13: shrinkage from no pooling to partial pooling")
axis(1, at = c(1, 2), labels = c("no pooling", "partial pooling"))
abline(h = grand, lty = 2, lwd = 1)
for (j in seq_along(sizes)) {
  segments(1, ybar[j], 2, pp[j], lwd = 2)
  points(1, ybar[j], pch = 19)
  points(2, pp[j], pch = 1, cex = 1.3)
  text(0.85, ybar[j], labels[j], cex = 0.9)
}
legend("topleft",
       legend = c("no-pooling mean", "partial-pooled mean", "overall mean"),
       pch = c(19, 1, NA), lty = c(NA, NA, 2), lwd = c(NA, NA, 1), bty = "n")
```

### Step 4 — Check the closed form by simulation

For one section, build the posterior for $\theta_j$ directly: with a Normal data model and a
Normal $N(\mu, \tau^2)$ level-two prior (known variances), the posterior for $\theta_j$ is Normal
with mean equal to the partial-pooled value and precision $n_j/\sigma^2 + 1/\tau^2$. Simulate from
that Normal and confirm the simulated mean matches the closed-form `pp` from Step 2. We use the
small section D, where shrinkage is largest.

```{r}
#| label: fig-lab13-posterior-D
#| fig-cap: "Simulated posterior for section D's mean (Normal with the closed-form partial-pooled mean and precision n/sigma^2 + 1/tau^2), with the closed-form mean and overall mean marked. The simulated mean lands on the closed-form value."
#| fig-alt: "A histogram of simulated posterior values for section D's mean, mound-shaped and centered near 80. A solid vertical line marks the closed-form partial-pooled mean on top of the histogram center, and a dashed vertical line further left marks the overall mean near 75, showing the section's estimate pulled toward but not all the way to the center."
jD <- 4                                            # section D
post_prec <- sizes[jD] / sigma^2 + 1 / tau^2       # posterior precision
post_sd   <- sqrt(1 / post_prec)
post_mean <- pp[jD]                                # = closed-form partial-pooled mean

set.seed(1313)
draws <- rnorm(50000, post_mean, post_sd)
hist(draws, breaks = 40, freq = FALSE, col = "grey85", border = "white",
     xlab = "section D mean score (posterior draws)",
     main = "Lab 13: section D posterior")
abline(v = post_mean, lwd = 2)
abline(v = grand, lwd = 2, lty = 2)
legend("topright",
       legend = c(sprintf("closed-form PP mean %.2f", post_mean),
                  sprintf("overall mean %.2f", grand)),
       lty = c(1, 2), lwd = 2, bty = "n")

c(sim_mean = mean(draws), closed_form = post_mean,
  abs_diff = abs(mean(draws) - post_mean))
```

### Illustrative — the same model with the course packages

The block below is **illustrative only — it runs with the course packages** (`rstanarm`,
`bayesrules`) and is **not executed here**. It shows the genuine hierarchical fit you would run once
those packages are installed; it estimates $\mu$, $\tau$, and $\sigma$ jointly and returns a
posterior for every section mean.

```r
# illustrative — runs with the course packages, not executed in this lab
library(rstanarm)
long <- data.frame(
  score   = unlist(scores),
  section = factor(rep(labels, sizes))
)
model <- stan_glmer(
  score ~ 1 + (1 | section),         # varying intercept = partial pooling
  data   = long,
  family = gaussian,
  seed   = 1313
)
# partial-pooled section means with credible intervals:
coef(model)$section
posterior_interval(model, prob = 0.95)
```

## Verify

Your work is on track if **all three** hold:

1. In Step 2, every `partial_pooled` value lies **between** that section's `no_pooling` value and the
   overall `complete_pooling` mean — partial pooling never overshoots either extreme.
2. The smallest section (D, $n = 6$) has the **smallest** `lambda` and moves the **most** toward the
   overall mean, while the largest section (B, $n = 30$) has the largest `lambda` and barely moves.
3. In Step 4, the simulated posterior mean matches the closed-form partial-pooled mean. (Concrete
   success criterion: `abs(mean(draws) - post_mean) < 0.1`.)

If any check fails, see the next section.

## When it breaks

- **`object 'pp' not found` in Step 3 or 4** — Step 2 did not run; chunks share state only when
  executed in sequence. **Fix** by rendering the whole `.qmd` top to bottom.
- **Numbers don't match the text** — confirm `set.seed(1313)` ran *in the same chunk* before the
  random draw; a missing or moved seed is the most common **error** here. **Troubleshoot** by
  re-rendering from a clean session.
- **A partial-pooled value lands *outside* the no-pooling/overall range** — you likely swapped
  `lambda` and `1 - lambda`; the weight `lambda` multiplies `ybar` (the section's own data), not the
  overall mean. Re-check Step 2.
- **If the Step 4 difference is large** — verify `post_prec` uses `sigma^2` and `tau^2` (variances,
  not sds) and that `jD <- 4`; a mismatched index compares the wrong section.

## AI use note

| Tool | Purpose | Verification |
|---|---|---|
| LLM assistant (e.g. Claude) | Explain an R error message or suggest a base-R plotting tweak for the shrinkage figure | Re-render the chunk yourself; confirm the table, figure, and the Step 4 difference satisfy the Verify criteria before trusting any AI-suggested edit. |

Disclose AI assistance per the syllabus. AI may help you *understand* code; the runnable result and
its interpretation must be your own and must pass the Verify checks.

## See also

- [Week 13 — Hierarchical models](../notes/week-13-hierarchical-models.qmd)
- [R + VS Code + Quarto setup](../resources/r-vscode-quarto-setup.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)

---

*The graded deliverable, its rubric, point values, and due date live in the LMS (Blackboard),
which is authoritative. None of those are posted here; this lab is the public, ungraded practice
path.*
