---
title: "Lab 9 — Bayesian regression"
subtitle: "Fit a line to synthetic study-hours/exam-score data and read coefficients with their uncertainty"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab is the hands-on companion to
> [Week 09 — Bayesian regression I](../notes/week-09-bayesian-regression-i.qmd). There you wrote
> the model $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ and saw that the posterior is a
> distribution over the coefficients. Here you *produce* a fitted line from synthetic data, read
> the slope and intercept, and — using simulation — turn the single line into a sense of the
> uncertainty around it.

## Goal

Fit a simple regression to synthetic study-hours/exam-score data, and read coefficient estimates
with their uncertainty. By the end you will have (1) a scatterplot with a fitted line, (2)
numeric estimates of intercept and slope, and (3) an interval around the slope, all reproduced
from a seeded, re-runnable `.qmd`.

## Setup

First, make sure your local toolchain is ready: R, VS Code, and Quarto, following the
[R + VS Code + Quarto setup page](../resources/r-vscode-quarto-setup.qmd). Then, create a new file
named `lab-09.qmd`, paste the chunks below in order, and render. Every chunk uses **base R only**
— no add-on packages — so it runs on a clean install. Seeds are fixed so your numbers match the
text.

## Steps

### Step 1 — Simulate the data

We invent a known truth — intercept $55$, slope $4$, noise sd $8$ — then draw 30 students. Keeping
the truth known lets you later check that your estimates recover it.

```{r}
#| label: lab09-simulate
set.seed(909)
n <- 30
x <- runif(n, 0, 10)                 # hours studied
y <- 55 + 4 * x + rnorm(n, 0, 8)     # exam score, with noise
head(round(cbind(hours = x, score = y), 1))
```

### Step 2 — Plot and fit the line

Fit the least-squares line with `lm()`. As Week 9 explained, this single line is the
**likelihood-dominated center** of the Bayesian posterior when priors are weak.

```{r}
#| label: fig-lab09-scatter
#| fig-cap: "Synthetic study-hours vs. exam-score data with the least-squares fit overlaid."
#| fig-alt: "Scatterplot of 30 grey points, hours studied on the x-axis from 0 to 10 and exam score on the y-axis, rising to the right; a solid black line passes through the cloud capturing the upward trend."
fit <- lm(y ~ x)
plot(x, y, pch = 19, col = "grey40",
     xlab = "Hours studied (x)", ylab = "Exam score (y)",
     main = "Lab 9: fitted regression line")
abline(fit, lwd = 2)
legend("topleft", legend = c("data", "fitted line"),
       pch = c(19, NA), lty = c(NA, 1), lwd = c(NA, 2),
       col = c("grey40", "black"), bty = "n")
coef(fit)
```

### Step 3 — Read the coefficients and a classical interval

`lm()` gives a point estimate per coefficient and a standard error. We can build a quick interval
for the slope to anchor intuition. (A full Bayesian credible interval comes from the posterior in
Step 4; here we just read what the data alone say.)

```{r}
#| label: lab09-coef-interval
round(coef(fit), 3)                   # intercept (b0) and slope (b1)
round(confint(fit, "x", level = 0.95), 3)   # classical 95% interval for the slope
```

### Step 4 — Approximate the slope's uncertainty by simulation

To *see* "a distribution over lines" without add-on packages, we use a simple resampling loop:
repeatedly resample the rows (with replacement), refit, and collect the slope. The spread of these
slopes mimics the posterior spread when the prior is weak — the simulation idea you met in Week 7,
now applied to a coefficient.

```{r}
#| label: fig-lab09-slope-dist
#| fig-cap: "Distribution of the fitted slope across 2000 resamples of the data; the spread visualizes coefficient uncertainty."
#| fig-alt: "Histogram of resampled slope estimates, roughly bell-shaped and centered near 4, with most mass between about 2 and 6, illustrating uncertainty in the slope."
set.seed(909)
B <- 2000
slopes <- numeric(B)
for (b in 1:B) {
  idx <- sample(n, n, replace = TRUE)
  slopes[b] <- coef(lm(y[idx] ~ x[idx]))[2]
}
hist(slopes, breaks = 30, col = "grey80", border = "white",
     xlab = "resampled slope estimate", main = "Lab 9: slope uncertainty")
abline(v = mean(slopes), lwd = 2)
legend("topright", legend = "mean of resampled slopes",
       lty = 1, lwd = 2, bty = "n")

c(mean_slope = mean(slopes),
  lower      = quantile(slopes, 0.025),
  upper      = quantile(slopes, 0.975))
```

Report the slope the way Week 9 insists: a **center with an interval**, e.g. "about 4 points per
hour, with a 95% interval of roughly $[\,\text{lower},\ \text{upper}\,]$."

### Illustrative — the same fit with the course packages

The block below is **illustrative only — it runs with the course packages** (`rstanarm`,
`bayesrules`) and is **not executed here**. It shows the genuine Bayesian fit you would run once
those packages are installed; it returns a posterior over the coefficients directly.

```r
# illustrative — runs with the course packages, not executed in this lab
library(rstanarm)
model <- stan_glm(
  y ~ x,
  data    = data.frame(x = x, y = y),
  family  = gaussian,
  prior_intercept = normal(55, 10),
  prior           = normal(4, 2),
  prior_aux       = exponential(1 / 8),
  seed    = 909
)
posterior_interval(model, prob = 0.95)   # credible intervals for b0 and b1
```

## Verify

Your work is on track if **all three** hold:

1. The fitted intercept and slope from Step 2 are near the *true* values **55** and **4** (within
   a point or two), confirming the data recover the truth.
2. In Step 4, the **mean of the resampled slopes** sits close to the `lm()` slope from Step 2 —
   the simulation center matches the closed-form least-squares estimate. (Concrete success
   criterion: `abs(mean(slopes) - coef(fit)[2]) < 0.3`.)
3. The 2.5%–97.5% resample interval for the slope **excludes 0**, matching the conclusion that
   study hours have a credible positive effect.

If any check fails, see the next section.

## When it breaks

- **`could not find function "lm"`** or a blank plot — you likely ran the chunk before R finished
  loading; **fix** by rendering the whole `.qmd` top to bottom so chunks execute in order.
- **Numbers don't match the text** — confirm the `set.seed(909)` line ran *in the same chunk*
  before the random draw; a missing or moved seed is the most common **error** here. **Troubleshoot**
  by re-rendering from a clean session.
- **`object 'fit' not found` in Step 4** — Step 2 did not run; chunks share state only when
  executed in sequence. Render the full document, not a single chunk.
- **If `confint()` fails**, your `lm` object is missing; re-run Step 2 first.

## AI use note

| Tool | Purpose | Verification |
|---|---|---|
| LLM assistant (e.g. Claude) | Explain an R error message or suggest a base-R plotting tweak | Re-render the chunk yourself; confirm the figure and printed numbers match the Verify criteria before trusting any AI-suggested edit. |

Disclose AI assistance per the syllabus. AI may help you *understand* code; the runnable result and
its interpretation must be your own and must pass the Verify checks.

## See also

- [Week 09 — Bayesian regression I](../notes/week-09-bayesian-regression-i.qmd)
- [R + VS Code + Quarto setup](../resources/r-vscode-quarto-setup.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)

---

*The graded deliverable, its rubric, point values, and due date live in the LMS (Blackboard),
which is authoritative. None of those are posted here; this lab is the public, ungraded practice
path.*
