---
title: "Lab 7 — Grid approximation in Quarto"
subtitle: "Build, sample, and verify a posterior on a grid, then render it as a reproducible report"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab puts [Week 07](../notes/week-07-simulation-first-computation.qmd) into your
> hands: you will approximate the bike-proportion posterior on a grid, draw from it, check the simulated
> mean against the closed-form Beta posterior, and assemble the whole thing as a seeded, reproducible
> Quarto report. Because this problem is conjugate, you can *verify* every approximation against a known
> truth — which is the habit the lab is really teaching.

## Goal

Approximate a posterior on a grid, draw from it, compare to the exact Beta, and assemble a reproducible
Quarto report. Concretely, by the end you will have a single `.qmd` that:

- builds a grid approximation of the bike-proportion posterior,
- samples posterior draws and summarizes them (mean and 95% credible interval),
- overlays the grid result on the exact $\text{Beta}(10,18)$ posterior, and
- renders cleanly to HTML with a session record.

## Setup

Confirm your local toolchain runs before you start: R must be discoverable by Quarto. If a render
complains it cannot find R, work through the [setup page](../resources/r-vscode-quarto-setup.qmd) first
(put R's `bin` on PATH, set `QUARTO_R`, or render from a shell that already has R on PATH). Everything
below is **base R only** — no add-on packages — so once R renders a trivial chunk, you are ready.

Create a new file `lab07.qmd` and copy the steps into it as chunks. Keep each chunk small so a failing
line is easy to find.

## Steps

### Step 1 — Lay down the grid and the model

The recurring problem: prior $\text{Beta}(2,2)$, data of **8 bikers in 24** surveyed (Binomial), whose
conjugate posterior is $\text{Beta}(10,18)$. Build the grid and evaluate prior $\times$ likelihood.

```{r}
#| label: setup-grid
theta <- seq(0, 1, length.out = 1001)         # 1001-point grid on [0,1]
prior <- dbeta(theta, 2, 2)                     # Beta(2,2) prior density
like  <- dbinom(8, size = 24, prob = theta)     # Binomial likelihood: y = 8, n = 24
unnorm <- like * prior                           # unnormalized posterior = prior x likelihood
post_grid <- unnorm / sum(unnorm)                # normalize: weights sum to 1
head(data.frame(theta, post_grid), 3)
```

### Step 2 — Overlay the grid posterior on the exact Beta

A correct grid approximation should sit on top of the exact $\text{Beta}(10,18)$ curve.

```{r}
#| label: fig-lab-overlay
#| fig-cap: "Grid approximation (points) overlaid on the exact Beta(10,18) posterior (curve) for the bike proportion. Agreement confirms the grid is built correctly."
#| fig-alt: "A smooth density curve for Beta(10,18) peaking just below 0.36, with open circles from the grid approximation lying directly on the curve across the whole range from about 0.15 to 0.65."
exact <- dbeta(theta, 10, 18)
grid_density <- post_grid / (theta[2] - theta[1])   # convert weights to density units
plot(theta, exact, type = "l", lwd = 2,
     xlab = expression(theta), ylab = "posterior density",
     main = "Grid vs exact Beta(10,18)")
points(theta[seq(1, 1001, by = 20)], grid_density[seq(1, 1001, by = 20)],
       pch = 1, cex = 0.7)
legend("topright", legend = c("exact Beta(10,18)", "grid approximation"),
       lty = c(1, NA), pch = c(NA, 1), lwd = c(2, NA), bty = "n")
```

### Step 3 — Sample posterior draws and summarize

Draw from the grid with probabilities equal to the normalized weights, then summarize like any data.
**Seed first** so the result reproduces.

```{r}
#| label: fig-lab-draws
#| fig-cap: "Histogram of 20,000 grid-sampled posterior draws with the simulated mean (solid) and 95 percent credible interval (dashed)."
#| fig-alt: "A bell-shaped histogram of posterior draws centered near 0.36 ranging from about 0.2 to 0.55, with a solid vertical line at the simulated mean near 0.357 and two dashed vertical lines marking the credible interval near 0.20 and 0.53."
set.seed(2026)
draws <- sample(theta, size = 20000, replace = TRUE, prob = post_grid)
sim_mean <- mean(draws)
ci <- quantile(draws, c(0.025, 0.975))
hist(draws, breaks = 40, freq = FALSE, col = "grey90", border = "white",
     xlab = expression(theta), main = "20,000 grid-sampled posterior draws")
abline(v = sim_mean, lwd = 2)
abline(v = ci, lty = 2)
c(sim_mean = sim_mean, lower = ci[1], upper = ci[2])
```

### Step 4 — Add a session record and render

End the report with a session record, then render the whole document.

```{r}
#| label: lab-session
sessionInfo()
```

Render from the terminal (with R on PATH) or VS Code's render button:

```bash
quarto render lab07.qmd
```

## Verify

Your approximation is correct if the **simulated mean matches the closed-form posterior mean**. The
closed form is $\dfrac{\alpha + y}{\alpha + \beta + n} = \dfrac{10}{28} \approx 0.357$. Check it:

```{r}
#| label: verify-check
closed_form <- 10/28
report <- c(closed_form = closed_form, simulated = sim_mean,
            abs_diff = abs(closed_form - sim_mean))
report
stopifnot(abs(closed_form - sim_mean) < 0.01)   # passes when grid + draws agree with the truth
```

Success criterion: the absolute difference is below 0.01, and the `stopifnot` does not error. If it
errors, your grid is too coarse, your draw count is too small, or you forgot to normalize — revisit
Step 1. (As a second check, your 95% **credible interval** should run roughly from about 0.20 to 0.53;
say "credible interval," never "confidence interval," for a posterior.)

## AI use note

- **Tool.** A general coding assistant in your editor.
- **Purpose.** Acceptable for scaffolding a chunk, decoding an R error, or drafting a caption — not for
  deciding what the posterior means.
- **Verification.** Re-run any AI-suggested code and confirm the verify step still passes
  (`|closed_form - simulated| < 0.01`) and every random chunk is seeded. Disclose AI use per the course
  AI policy in the LMS. Unverifiable AI output does not go in a report you submit.

The graded deliverable, its rubric, point values, and due date live in the **LMS** — this lab page does
not contain any of those. For grading specifics, the LMS (Blackboard) is authoritative.

## See also

- [Week 07 — Simulation-first computation](../notes/week-07-simulation-first-computation.qmd)
- [R · VS Code · Quarto setup](../resources/r-vscode-quarto-setup.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)
- [Syllabus](../syllabus.qmd)
