---
title: "Lab 4 — Beta-Binomial simulation"
subtitle: "Simulate, summarize, and read a Beta posterior for the bike-to-campus proportion"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
toc: true
---

> **Purpose.** This lab puts the Week 4 derivation to work. You will simulate the Beta posterior
> for the bike-to-campus proportion, summarize it with a posterior mean and a 95% credible
> interval, and confirm the simulation agrees with the closed-form answer. Read it alongside
> [Week 04 — The Beta-Binomial model](../notes/week-04-beta-binomial.qmd).

## The idea

The core concept is that the closed-form posterior and a simulated posterior are two views of the
same object. Week 4 derived that a Beta(2,2) prior plus 8 bikers out of 24 gives a posterior
Beta(10,18) with mean $10/28 \approx 0.357$. Here we *also* obtain that posterior by drawing
random values with `rbeta`, summarizing the draws, and checking that the simulated summaries match
the formulas. Simulation matters because most real models later in the course have **no** clean
closed form — we will only have draws — so learning to trust and read draws now pays off later.

## Goal

Simulate a Beta posterior for the bike-to-campus proportion, summarize it (mean + 95% credible
interval), and read the result — confirming the simulated mean matches the closed-form posterior
mean $10/28 \approx 0.357$.

## Setup

First, make sure your local environment renders. If R, VS Code, or Quarto are not yet working,
follow the [R + VS Code + Quarto setup page](../resources/r-vscode-quarto-setup.qmd) before you
start. Then, create a new `.qmd` file in your course folder (for example
`lab-04-mywork.qmd`), paste the chunks below one at a time, and render after each so you catch
errors early. Everything here uses **base R only** — no add-on packages — so there is nothing
extra to install.

## Steps

### Step 1 — Set the model and the closed-form answer

We hard-code the recurring numbers and compute the closed-form posterior so we have a target to
hit. Seeding is not needed yet (no randomness), but we set it now so the whole document is
reproducible.

```{r}
#| label: setup-numbers
set.seed(404)

# Prior Beta(alpha, beta) and data
alpha <- 2; beta <- 2
y <- 8; n <- 24            # 8 bikers of 24 respondents

# Conjugate update: successes onto alpha, failures onto beta
a_post <- alpha + y        # 10
b_post <- beta + (n - y)   # 18

# Closed-form summaries
post_mean_cf <- a_post / (a_post + b_post)
ci_cf <- qbeta(c(0.025, 0.975), a_post, b_post)

c(a_post = a_post, b_post = b_post,
  post_mean_cf = post_mean_cf,
  ci_lo = ci_cf[1], ci_hi = ci_cf[2])
```

You should see `a_post = 10`, `b_post = 18`, and a posterior mean near `0.357`.

### Step 2 — Simulate the posterior

Now draw many values from the posterior Beta and summarize the draws. The simulated mean and the
empirical 2.5%/97.5% quantiles should land on the closed-form values from Step 1.

```{r}
#| label: simulate
draws <- rbeta(50000, a_post, b_post)

post_mean_sim <- mean(draws)
ci_sim <- quantile(draws, c(0.025, 0.975))

c(post_mean_sim = post_mean_sim,
  ci_lo_sim = ci_sim[1], ci_hi_sim = ci_sim[2])
```

### Step 3 — Plot the simulated posterior with its credible interval

This figure is the deliverable's centerpiece: a histogram of the draws with the 95% credible
interval shaded and the posterior mean marked.

```{r}
#| label: fig-lab-posterior-ci
#| fig-cap: "Simulated posterior for the bike-to-campus proportion (50,000 draws from Beta(10,18)) with the equal-tailed 95% credible interval shaded and the posterior mean marked."
#| fig-alt: "A density histogram of simulated bike-proportion values, mound-shaped and centered near 0.36. The central 95 percent of the mass is shaded between two dashed vertical lines at roughly 0.20 and 0.53, and a solid vertical line marks the posterior mean near 0.357."
h <- hist(draws, breaks = 40, plot = FALSE)
plot(h, freq = FALSE, col = "white", border = "grey60",
     xlab = expression(theta ~ "(proportion who bike)"),
     ylab = "density",
     main = "Lab 4: simulated posterior with 95% credible interval")
in_ci <- h$mids >= ci_sim[1] & h$mids <= ci_sim[2]
rect(h$breaks[-length(h$breaks)][in_ci], 0,
     h$breaks[-1][in_ci], h$density[in_ci],
     col = "grey80", border = "grey60")
abline(v = ci_sim, lwd = 2, lty = 2)
abline(v = post_mean_sim, lwd = 2, lty = 1)
legend("topright",
       legend = c(sprintf("95%% CI [%.3f, %.3f]", ci_sim[1], ci_sim[2]),
                  sprintf("mean %.3f", post_mean_sim)),
       lty = c(2, 1), lwd = 2, bty = "n")
```

### Step 4 — Overlay the exact density (optional check)

Adding the exact Beta(10,18) curve over the simulated density is a strong visual confirmation that
the draws reproduce the closed-form posterior.

```{r}
#| label: fig-lab-overlay
#| fig-cap: "Simulated posterior density (histogram) for the bike proportion with the exact Beta(10,18) density overlaid as a solid curve; the two agree closely."
#| fig-alt: "The same mound-shaped histogram of simulated bike-proportion values, with a smooth solid curve for the exact Beta(10,18) density drawn on top. The curve traces the top of the histogram bars almost exactly, showing the simulation matches the closed-form posterior."
plot(h, freq = FALSE, col = "white", border = "grey70",
     xlab = expression(theta ~ "(proportion who bike)"),
     ylab = "density",
     main = "Simulated vs exact posterior")
xx <- seq(0, 1, length.out = 400)
lines(xx, dbeta(xx, a_post, b_post), lwd = 2)
legend("topright", legend = "exact Beta(10,18)", lwd = 2, bty = "n")
```

### Reproducible workflow

Keep the work tidy so it renders the same way for anyone (including future you):

- Put all of this in a single `.qmd` **file** with a clear name (e.g. `lab-04-mywork.qmd`).
- Call `set.seed(404)` once near the top so the simulation is reproducible on every **render**.
- **Save** the file, then **render** the whole document end to end (not just chunk-by-chunk) to
  confirm it builds clean from a fresh state.
- Use base R only; do not add library calls, so the `.qmd` renders on any machine with R.

### When it breaks

A few common snags and fixes:

- **Render error: "could not find function rbeta".** You are almost certainly outside an R chunk
  or R is not on PATH. Confirm the chunk fence is `{r}` and recheck the setup page.
- **The shaded interval looks wrong / mirrored.** You likely swapped the Beta shapes. Order is
  `rbeta(n, shape1, shape2)` = `rbeta(n, a_post, b_post)`; if `a_post` and `b_post` are reversed,
  the distribution flips. Troubleshoot by printing `c(a_post, b_post)` and confirming `10, 18`.
- **Simulated mean is noticeably off from 0.357.** Either the seed/draw count changed or the data
  inputs are wrong. Increase draws to 50,000 and re-verify `y = 8`, `n = 24`. If a chunk fails,
  comment out everything below it, render, then re-add chunks one at a time to isolate the break.

### A hands-on step

For the hands-on step, change one input and predict the result *before* you run it: set `y <- 12`
(keeping `n <- 24`), re-render, and check whether the posterior mean moved up toward `0.5` as you
expected. This is the concrete activity carried into Lab 4 on the LMS.

## Verify

Your lab succeeds if, in one render, **all** of the following hold:

- The simulated posterior mean (`post_mean_sim`) agrees with the closed-form mean
  `10/28 = 0.357...` to about two decimal places.
- The simulated 95% interval (`ci_sim`) agrees with `qbeta(c(0.025, 0.975), 10, 18)` to about two
  decimal places.
- The overlay in Step 4 shows the exact Beta(10,18) curve tracing the top of the histogram.

A compact numerical check you can include:

```{r}
#| label: verify-check
data.frame(
  quantity = c("posterior mean", "CI lower", "CI upper"),
  closed_form = c(post_mean_cf, ci_cf[1], ci_cf[2]),
  simulated   = c(post_mean_sim, ci_sim[1], ci_sim[2]),
  abs_diff    = abs(c(post_mean_cf, ci_cf[1], ci_cf[2]) -
                    c(post_mean_sim, ci_sim[1], ci_sim[2]))
)
```

All `abs_diff` values should be small (well under 0.01).

## AI use note

| Tool | Purpose | Verification |
|---|---|---|
| LLM assistant (optional) | explain an R error message or suggest a base-R idiom | Re-run the chunk yourself; confirm the simulated mean still matches `10/28` and the figure renders. Never paste a result you did not reproduce. |

Use of an AI assistant is optional and never a substitute for running the code. Always verify any
suggested code against the closed-form posterior before keeping it.

## See also

- [Week 04 — The Beta-Binomial model](../notes/week-04-beta-binomial.qmd)
- [R + VS Code + Quarto setup](../resources/r-vscode-quarto-setup.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)

The graded deliverable, its rubric, point values, and the due date live in the LMS (Blackboard) —
this page is study and practice only and posts none of those.
