---
title: "Week 6 — Posterior predictive thinking"
subtitle: "Predicting new data and checking whether the model could have produced ours"
page_mode: normal_week
week_contract: public_note
source_mode: source-informed
readings:
  - "Bayes Rules! Ch 8 — Posterior inference & prediction — supports the move from inference about θ to prediction of new data and posterior predictive checking"
toc: true
---

## The week question

We spent five weeks asking *what does the data tell us about the parameter?* This week we turn the
telescope around and ask a different question: **if the model is right, what new data should we expect
to see — and does the data we actually saw look like something this model could have produced?**

## Where we are and why this matters

By now the posterior is an old friend. In the running bike-to-campus survey we started with a mild
prior $f(\theta) = \text{Beta}(2,2)$ on $\theta$, the unknown proportion of students who bike to
campus, observed $y = 8$ bikers out of $n = 24$ surveyed, and updated to the posterior
$f(\theta \mid y) = \text{Beta}(10, 18)$, with posterior mean $10/28 \approx 0.357$. That posterior
is a full distribution over $\theta$ — a statement about a quantity we will never directly observe.
You cannot walk onto campus and *measure* "the proportion." You can only ever observe **counts in
samples**.

That gap is the whole point of this week. Inference lives in $\theta$-space; the world hands us data
in $y$-space. To predict the next survey, or to ask whether our model is any good, we have to push
our posterior uncertainty about $\theta$ back out into uncertainty about *future data*. The tool that
does this is the **posterior predictive distribution**, and the discipline it enables —
**posterior predictive checking** — is one of the most practical, honest things Bayesian modeling
offers: a built-in way to ask the model "could you have generated what I actually saw?"

This connects two ideas you already hold. First, the posterior is over $\theta$, not over data
(Weeks 3–5). Second, every model carries a **data model** — the likelihood's distributional form,
e.g. Binomial counts (Week 4). Prediction is what you get when you let those two talk to each other.

## Learning goals

By the end of this week you should be able to:

- Define the **posterior predictive distribution** as the data model averaged over the posterior.
- Explain *why* prediction integrates out $\theta$, and what that averaging buys you.
- Distinguish **parameter uncertainty** from **predictive uncertainty**, and say which is wider and why.
- Simulate posterior predictive data and read a predictive distribution and predictive interval.
- Run a **posterior predictive check**: compare simulated replicated data to the observed data.
- Interpret a passed or failed check carefully — what it does, and does not, license you to conclude.

## Core vocabulary

- **Data model / likelihood form** — the distribution we assume generated each observation given
  $\theta$ (here, Binomial counts). It is the recipe for turning a parameter into data.
- **Posterior predictive distribution** — written $f(y_{\text{new}} \mid y)$: the distribution of a
  *new* dataset, given the data we already have, after accounting for all our uncertainty about $\theta$.
- **Replicated data** ($y^{\text{rep}}$) — synthetic datasets generated from the model as if it were
  true, used for checking. "Replicated" means *the same kind and size of data we collected*.
- **Predictive uncertainty** — spread in $y_{\text{new}}$. It bundles together two sources: we are
  unsure about $\theta$, **and** even a known $\theta$ produces random data.
- **Posterior predictive check (PPC)** — comparing observed data (or a summary of it) to the
  distribution of replicated data, to judge model adequacy.

## From "what is θ?" to "what data comes next?"

Suppose we knew $\theta$ exactly — say $\theta = 0.357$. Predicting the number of bikers in a fresh
sample of $m$ students would be a pure probability problem: $y_{\text{new}} \sim \text{Binomial}(m,
0.357)$. There would still be randomness, because sampling is random, but it would be the *only*
randomness.

We do not know $\theta$ exactly. We have a posterior $f(\theta \mid y)$ that spreads belief across a
range of plausible values. The honest prediction must account for *both* the randomness in sampling
**and** our remaining uncertainty about which $\theta$ is true. The posterior predictive distribution
does exactly this by **averaging the data model over the posterior** — integrating out $\theta$:

$$
f(y_{\text{new}} \mid y) \;=\; \int f(y_{\text{new}} \mid \theta)\; f(\theta \mid y)\; d\theta .
$$

Read this left to right as a recipe, not just a formula. For every possible value of $\theta$, take
the data model $f(y_{\text{new}} \mid \theta)$ — what data that $\theta$ would produce — and weight it
by how much the posterior believes in that $\theta$. Sum (integrate) over all $\theta$. What comes out
is a distribution over **future data** that already has our parameter uncertainty baked in.

There is a beautifully simple way to *do* this without integrals, and it is the way we will actually
work all term: **simulate**. Draw a $\theta$ from the posterior, then draw a dataset from the data
model at that $\theta$; repeat thousands of times. The collected datasets are draws from the posterior
predictive distribution. This two-step draw — first $\theta$, then $y_{\text{new}} \mid \theta$ — is
the computational heart of next week, and it is why prediction and simulation are taught back to back.

## Why predictive uncertainty is wider than parameter uncertainty

This is the idea most worth carrying out of the week. The posterior $f(\theta \mid y)$ describes our
uncertainty about a *number*. The posterior predictive $f(y_{\text{new}} \mid y)$ describes our
uncertainty about *future data*. The second is always at least as spread out as the first, and
usually wider, because it stacks **two** sources of variation:

1. **Parameter uncertainty** — we are not sure which $\theta$ is true (the width of the posterior).
2. **Sampling variability** — even if we knew $\theta$ perfectly, a fresh sample of size $m$ would
   come out differently each time (the spread of $\text{Binomial}(m, \theta)$).

Predictive uncertainty is the *sum* of these effects, not just the first. A common and costly error
is to predict using only the posterior — for example, predicting that a new sample of 24 will contain
"about $24 \times 0.357 \approx 8.6$ bikers, give or take a little" using only how tightly the
posterior is concentrated. That undersells the spread badly, because it forgets that sampling alone
scatters counts. We will see this concretely in the figures below.

## Posterior predictive checking: can the model generate data like ours?

A model is a story about how data came to be. A good story should be able to produce data that
resembles what we actually collected. **Posterior predictive checking** turns this into a procedure:

1. Fit the model and get the posterior.
2. Generate many **replicated datasets** $y^{\text{rep}}$ from the posterior predictive distribution —
   each the same size and shape as your real data.
3. Compute a **summary** (a "test quantity") on each replicate and on the observed data.
4. Compare: does the observed summary look like a typical draw from the replicates, or does it sit far
   out in the tail?

If the observed data looks like an ordinary member of the replicated crowd, the model has *passed*
this particular check — it is at least capable of producing data like ours. If the observed summary
is extreme relative to the replicates, the check has *flagged* a mismatch: the model, as specified,
would be surprised by your own data, which is a signal that some assumption may be off.

The choice of summary matters. You check what you care about: a proportion, a maximum, a count of
zeros, the spread, the number of "extreme" observations. A model can pass on the mean and fail badly
on the variance. Predictive checking is therefore less a single pass/fail stamp than a flexible
diagnostic conversation with the model.

### Worked example — predicting bikers in a NEW sample (recurring case)

Take our posterior $f(\theta \mid y) = \text{Beta}(10, 18)$ from the bike survey. We plan a **new**
survey of $m = 24$ students and ask: how many bikers should we expect, *honestly* accounting for all
our uncertainty? We simulate the posterior predictive distribution with the two-step draw, then
overlay what we would have gotten if we (wrongly) pretended $\theta$ were fixed at its posterior mean.

```{r}
#| label: fig-bike-ppd
#| fig-cap: "Posterior predictive distribution for the number of bikers in a new sample of m = 24, from the Beta(10,18) posterior (gray), compared with the narrower Binomial(24, 0.357) you would wrongly get by fixing θ at its posterior mean (outline). Predictive spread is the wider one."
#| fig-alt: "Two overlaid histograms of biker counts from 0 to 24. The gray filled histogram from the full posterior predictive simulation is centered near 8 or 9 and visibly wider; the outlined histogram from fixing theta at 0.357 is centered the same place but noticeably narrower, showing that ignoring parameter uncertainty understates predictive spread."
set.seed(606)
m <- 24
N <- 20000

# Two-step draw: first theta from the posterior, then data given theta
theta_draws <- rbeta(N, 10, 18)              # posterior Beta(10,18)
y_ppd       <- rbinom(N, size = m, prob = theta_draws)  # new counts

# The WRONG, too-narrow version: fix theta at the posterior mean
theta_mean  <- 10 / 28
y_fixed     <- rbinom(N, size = m, prob = theta_mean)

brks <- seq(-0.5, m + 0.5, by = 1)
h1 <- hist(y_ppd,   breaks = brks, plot = FALSE)
h2 <- hist(y_fixed, breaks = brks, plot = FALSE)

plot(h1, freq = FALSE, col = "grey80", border = "grey50",
     main = "", xlab = "number of bikers in a new sample of 24",
     ylab = "predictive probability",
     ylim = c(0, max(h1$density, h2$density) * 1.05))
plot(h2, freq = FALSE, col = NA, border = "black", lwd = 2, add = TRUE)
abline(v = m * theta_mean, lty = 2)
legend("topright", bty = "n",
       legend = c("posterior predictive (honest)",
                  "fixed-theta (too narrow)",
                  "posterior-mean count"),
       fill = c("grey80", NA, NA),
       border = c("grey50", NA, NA),
       lty = c(NA, 1, 2), lwd = c(NA, 2, 1))
```

Both distributions center near $24 \times 0.357 \approx 8.6$. But the gray posterior predictive is
**wider** — that extra width is exactly the parameter uncertainty the fixed-$\theta$ version threw
away. A defensible "central guess plus interval" comes from the gray one. We can read a predictive
interval straight off the simulation:

```{r}
#| label: fig-bike-ppi
#| fig-cap: "An approximate 90% posterior predictive interval for the number of bikers in a new sample of 24, read off the simulated posterior predictive counts as the 5th and 95th percentiles."
#| fig-alt: "Histogram of simulated biker counts for a new sample of 24, with two vertical dashed lines marking the 5th and 95th percentiles that bracket the central 90 percent of predicted counts, roughly spanning the mid-single-digits up to the low-teens."
set.seed(606)
m <- 24
theta_draws <- rbeta(20000, 10, 18)
y_ppd <- rbinom(20000, size = m, prob = theta_draws)

pi90 <- quantile(y_ppd, c(0.05, 0.95))
hist(y_ppd, breaks = seq(-0.5, m + 0.5, by = 1), freq = FALSE,
     col = "grey85", border = "grey50",
     main = "", xlab = "number of bikers in a new sample of 24",
     ylab = "predictive probability")
abline(v = pi90, lty = 2, lwd = 2)
legend("topright", bty = "n", lty = 2, lwd = 2,
       legend = sprintf("approx 90%% predictive interval: [%d, %d]",
                        round(pi90[1]), round(pi90[2])))
```

The interpretation is plain-language and honest: *if our model is right, then in a new sample of 24
students we would predict roughly the mid-single-digits to low-teens bikers, with about 90%
predictive probability.* Note this is a **predictive** interval about future *counts*, not a credible
interval about $\theta$ — different objects, different widths. Always say which one you mean.

### Worked example — a posterior predictive CHECK on the bike data (recurring case)

Now the checking move. We have the observed data: $y = 8$ bikers in $n = 24$. Is that an ordinary
thing for our fitted model to produce? We generate many **replicated** datasets of the same size
($n = 24$) from the posterior predictive distribution, summarize each by its observed proportion
$y^{\text{rep}}/24$, and see where our actual observed proportion $8/24 \approx 0.333$ falls in the crowd.

```{r}
#| label: fig-bike-ppc
#| fig-cap: "Posterior predictive check for the bike survey: simulated replicated proportions (gray) from samples of size 24, with the observed proportion 8/24 marked. The observed value sits comfortably inside the bulk, so the model passes this check."
#| fig-alt: "Histogram of replicated sample proportions ranging from about 0.1 to 0.6, centered near 0.36, with a solid vertical line at the observed proportion of about 0.33 that lands near the middle of the distribution rather than in a tail, indicating the model could plausibly have produced the observed data."
set.seed(606)
n <- 24
theta_draws <- rbeta(10000, 10, 18)
y_rep <- rbinom(10000, size = n, prob = theta_draws)
prop_rep <- y_rep / n

obs_prop <- 8 / 24
hist(prop_rep, breaks = 20, freq = FALSE,
     col = "grey85", border = "grey60",
     main = "", xlab = "replicated proportion of bikers (samples of 24)",
     ylab = "density")
abline(v = obs_prop, lwd = 3)
legend("topright", bty = "n", lwd = 3,
       legend = sprintf("observed proportion = %.3f", obs_prop))

# A rough "predictive position": fraction of replicates at or below the observed
mean(prop_rep <= obs_prop)
```

The observed proportion lands near the middle of the replicated cloud — not out in a tail. The
printed fraction of replicates at or below the observed value is comfortably away from 0 and 1. This
check **passes**: a Beta-Binomial model with this posterior is entirely capable of generating data
like ours. That is reassuring but modest — it confirms consistency on *this summary*, nothing more.

(Of course this check should pass: we fit the model *to* this data, so the data ought to look typical
of its own posterior predictive. Predictive checks earn their keep on summaries the model was **not**
tuned to match — overdispersion, zero counts, extreme values — and that is the real diagnostic skill.)

### Worked example — transfer: a clinic wait-time count check (Gamma-Poisson)

A different context, same logic. A small campus clinic models the number of walk-ins per hour as
$\text{Poisson}(\lambda)$ and has arrived at a posterior $f(\lambda \mid y) = \text{Gamma}(40, 8)$
(the Gamma–Poisson conjugate update — a Gamma prior with Poisson counts, the same
posterior $\propto$ likelihood $\times$ prior logic in a new family — which we take as given for this
check; posterior mean $40/8 = 5$ walk-ins/hour). Staff suspect the real data has
*more* very-busy hours than a plain Poisson allows. A natural test quantity is the **maximum** count
seen in a 12-hour day. We generate replicated days and compare the observed maximum (say it was 14)
to the replicated maxima.

```{r}
#| label: fig-clinic-ppc
#| fig-cap: "Posterior predictive check for clinic walk-ins: the distribution of the maximum hourly count over a 12-hour replicated day from a Gamma(40,8)-Poisson model, with an observed maximum of 14 marked far in the right tail — a flagged mismatch."
#| fig-alt: "Histogram of the maximum hourly walk-in count per simulated 12-hour day, mostly between about 7 and 12, with a solid vertical line at 14 sitting in the extreme right tail beyond almost all replicated values, indicating the Poisson model rarely produces a day this busy and the check has flagged a problem."
set.seed(606)
hours <- 12
Nday  <- 5000

lambda_draws <- rgamma(Nday, shape = 40, rate = 8)   # posterior Gamma(40,8)
# For each replicated day, simulate 12 hourly counts and take the max
rep_max <- sapply(lambda_draws, function(lam) max(rpois(hours, lam)))

obs_max <- 14
hist(rep_max, breaks = seq(min(rep_max) - 0.5, max(c(rep_max, obs_max)) + 0.5, by = 1),
     freq = FALSE, col = "grey85", border = "grey60",
     main = "", xlab = "maximum hourly walk-in count in a 12-hour day",
     ylab = "predictive probability")
abline(v = obs_max, lwd = 3)
legend("topright", bty = "n", lwd = 3,
       legend = sprintf("observed maximum = %d", obs_max))

mean(rep_max >= obs_max)   # how often replicates are this extreme
```

Here the observed maximum sits far in the right tail, and the printed fraction of replicates that are
this extreme is tiny. The check has **flagged** a mismatch: a plain Poisson with this posterior almost
never produces a day this busy, yet we saw one. That does not tell us *what* to do — but it points
exactly where to look (perhaps the count is overdispersed, or busy hours cluster), which is what a
good check is for.

## A common mistake

**Treating parameter uncertainty as if it were predictive uncertainty — and getting the width wrong.**

The trap sounds reasonable: "My posterior says $\theta$ is about $0.357$ and fairly precise, so the
number of bikers in the next 24 should be about $8.6$, very tightly." That confuses two different
questions. The posterior's tightness is about the *parameter*; the next sample's spread also includes
**sampling variability**, which is often the larger part for a single new dataset. A predictive
statement built only from the posterior is **too confident** — its interval is too narrow, and it
will be "surprised" by ordinary future data far too often.

How to catch it: ask yourself *what object am I describing — a parameter or future data?* If it is
future data, you must simulate the full two-step draw (posterior $\theta$, then data $\mid \theta$),
or your interval is missing a whole source of variation. The figure `@fig-bike-ppd` is the antidote:
the wider gray distribution is honest; the narrow outline is the mistake.

## Interpretation guidance

What a **passed** check means: the model is *capable* of producing data like yours on the summary you
chose. It is a consistency check, not a certificate of truth. Many wrong models pass many checks; a
pass simply means you have not found *this* particular flaw. Passing one summary says nothing about
others you did not test.

What a **failed** check means and does *not* mean: a failure says the observed data is surprising
*under this model on this summary* — a genuine, useful signal that some assumption may be inadequate.
It does **not** tell you which assumption is wrong, it does **not** prove the model is "false" in any
absolute sense, and it does **not** by itself tell you the fix. It is a flashlight pointing at a
discrepancy, not a verdict. A sensible response is to look at *which* summary failed and reason about
what modeling change would address it (a different data model, more structure), then re-check.

Two framing reminders carried from earlier weeks. Predictive intervals are about **future data**, not
about $\theta$ — keep them verbally distinct from credible intervals. And as always, report a central
predictive value **with** its predictive interval; a single predicted number with no spread hides the
very uncertainty this week is built to expose.

## Practice (ungraded)

Use these to check your own understanding. No solutions are posted here.

1. In your own words, why does the posterior predictive distribution integrate out $\theta$ rather
   than just plug in the posterior mean? What would be lost by plugging in?
2. Using the bike posterior $\text{Beta}(10,18)$, sketch (or simulate) the posterior predictive for a
   new sample of $m = 50$. Would you expect the *proportion* spread to be wider or narrower than for
   $m = 24$? Why?
3. A classmate reports a "95% interval of $0.34$ to $0.37$ for the number of bikers next week."
   Name what has gone wrong and what they were probably computing instead.
4. For the clinic example, suggest *two different* test quantities (besides the maximum) you could
   use in a posterior predictive check, and say what kind of model failure each would be sensitive to.
5. A posterior predictive check passes on the mean but fails on the variance. What can and cannot you
   conclude, and what would you investigate next?

## Reading guide

This week maps to **Bayes Rules! Ch 8 — Posterior inference & prediction**. Chapter 8 develops three
strands of using a posterior: summarizing it, testing claims about $\theta$, and — our focus —
**prediction**. Read the chapter's treatment of the posterior predictive model as the bridge from
$\theta$-inference to data-prediction: its core message is the same averaging-over-the-posterior idea
we wrote as the integral above, and its simulation recipe is the same two-step draw we used in every
figure. As you read, hold our running bike example beside the book's worked case so the *structure*
transfers even though the numbers differ. The chapter also previews why simulation is the practical
engine for all of this, which is exactly where Week 7 goes. We adapt the chapter's concepts in the
course's own framing and synthetic examples; the text is used under CC BY-NC-SA 4.0.

## Public vs. graded

These notes and the ungraded practice above are public and for learning only. **No answer keys are
posted here**, and nothing on this page sets graded prompts, rubric values, point values, or due
dates. For anything that counts toward your grade — homework specifics, quiz scope, deadlines —
**the LMS (Blackboard) is authoritative**. Where this page and the LMS differ, the LMS governs.

## Looking ahead

Every figure this week leaned on the same trick — *draw a $\theta$, then draw data* — done thousands
of times. Next week, **Week 7 — Simulation-first computation**, we make that trick the explicit
foundation: building, summarizing, and trusting posteriors and predictions through simulation, and
turning it into a reproducible workflow. See [Week 7 — Simulation-first computation](../notes/week-07-simulation-first-computation.qmd).

## See also

- [Labs](../labs/index.qmd) — the reproducible computation strand; the predictive simulation here is
  built on the same two-step draw practiced in [Lab 7 — Grid approximation in Quarto](../labs/lab-07-grid-approximation-quarto.qmd).
- [Notation glossary](../resources/notation-glossary.qmd) — symbols for $\theta$, $y$,
  $f(\theta \mid y)$, and the posterior predictive $f(y_{\text{new}} \mid y)$.
- [Syllabus](../syllabus.qmd) — course shape and outcomes (O3, O8 are the focus this week).
