---
title: "Week 10 — Bayesian regression II"
subtitle: "More predictors, binary outcomes, and prediction that carries its uncertainty"
page_mode: math_derivation_week
week_contract: math_derivation
source_mode: source-informed
readings:
  - "Bayes Rules! Ch 11 — Extending simple regression to several predictors — supports the multiple-predictor model statement."
  - "Bayes Rules! Ch 12 — Poisson and negative-binomial regression — supports the transfer case for count outcomes."
  - "Bayes Rules! Ch 13 — Logistic regression — supports the binary-outcome model and the log-odds link."
toc: true
---

## Mathematical goal

Week 9 fit a single straight line to a continuous outcome. This week the goal is to write down — and
read off — two extensions of that model that are not optional decorations but the workhorses of applied
Bayesian modeling. First, we extend the linear predictor from one slope to **several**: a coefficient
for each of several predictors, each interpreted *holding the others fixed*. Second, we change the
**outcome type**: when the response is binary (yes/no, success/failure) a straight line through it is
nonsensical, so we model the **log-odds** of success as linear in the predictors — logistic regression.
We will state both models precisely, derive how a single logistic coefficient becomes an **odds ratio**
on the multiplicative scale, and insist throughout that any *prediction* we produce comes with a
posterior — a distribution, not a number.

The mathematical objects do not change from earlier weeks. We still have a prior $f(\theta)$ over
parameters $\theta$, a likelihood $L(\theta \mid y)$ from the data model, and a posterior
$f(\theta \mid y) \propto L(\theta \mid y)\, f(\theta)$. What changes is the *shape* of $\theta$ (now a
vector of coefficients) and the *form* of the data model (Bernoulli through a link, rather than Normal).

## Notation

This week introduces vector-valued parameters and a link function. The notation below extends the fixed
course conventions; nothing here overrides them.

| Symbol | Reads as | Note |
|---|---|---|
| $y_i$ | the outcome for observation $i$ | binary ($0/1$) in logistic regression |
| $x_{i1}, x_{i2}, \dots, x_{ik}$ | the $k$ predictor values for observation $i$ | one column per predictor |
| $\beta_0$ | the intercept coefficient | a parameter, so it has a prior |
| $\beta_1, \dots, \beta_k$ | the slope coefficients | one per predictor; each a parameter |
| $\pi_i$ | $P(y_i = 1)$, the success probability for observation $i$ | lives in $(0,1)$ |
| $\eta_i$ | the **linear predictor** $\beta_0 + \beta_1 x_{i1} + \cdots$ | lives on the whole real line |
| $\theta$ | the full parameter vector $(\beta_0, \beta_1, \dots, \beta_k)$ | (plus the noise sd in the Normal case) |
| $\operatorname{odds}(\pi)$ | $\pi / (1-\pi)$ | the odds of success |
| $\operatorname{logit}(\pi)$ | $\log\!\big(\pi/(1-\pi)\big)$ | the log-odds; the logistic **link** |

The prior is $f(\theta)$, the likelihood $L(\theta \mid y)$ is a function of $\theta$ (not a
distribution over it), and the posterior is $f(\theta \mid y)$. The symbol $\propto$ continues to mean
"equal up to a constant that does not depend on $\theta$"; what it drops is the marginal $f(y)$.

## Conceptual setup

Recall the Week 9 setup: $y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$ with $\varepsilon_i \sim
N(0,\sigma^2)$, priors on $\beta_0,\beta_1,\sigma$, and a posterior over the coefficients. Two
assumptions in that line are worth naming because this week relaxes or replaces them.

The first assumption is that **one predictor** is enough. Almost never true. If exam score depends on
both study hours *and* prior GPA, a model with only study hours forces the GPA effect to hide inside the
study-hours slope, biasing it. The fix is honest: give each predictor its own coefficient. The linear
predictor becomes a sum, and we *assume* (to start) that effects are **additive** — each predictor
contributes its slope times its value, and the coefficients add. We also assume each slope is a
*partial* effect: the change in the expected outcome for a one-unit change in that predictor **with the
other predictors held fixed**. That conditioning phrase is the whole interpretive content of multiple
regression; drop it and you will misread every coefficient.

The second assumption is that the outcome is continuous and a straight line through it is meaningful.
For a **binary** outcome $y_i \in \{0,1\}$ this fails. A line $\beta_0 + \beta_1 x_i$ ranges over the
whole real line, but a probability must sit in $(0,1)$; a linear model will happily predict $1.3$ or
$-0.2$. The remedy is to keep the linear predictor on the real line but **link** it to the probability
through a squashing function. We model the **log-odds** as linear:
$\operatorname{logit}(\pi_i) = \eta_i$, which means $\pi_i = 1/(1+e^{-\eta_i})$ always lands in
$(0,1)$. The data model is then Bernoulli: $y_i \mid \pi_i \sim \text{Bernoulli}(\pi_i)$.

Crucially, the Bayesian object is still a *posterior over coefficients*. Because each draw of
$(\beta_0,\beta_1,\dots)$ from the posterior implies a whole fitted curve, the Bayesian view of a
logistic fit is not one S-curve but a **posterior bundle of S-curves** — and any predicted probability
inherits that spread. Hold onto this: prediction is a distribution.

## The multiple-predictor model statement

With $k$ predictors and a continuous outcome, the model is

$$
y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots + \beta_k x_{ik} + \varepsilon_i,
\qquad \varepsilon_i \sim N(0, \sigma^2),
$$

with independent priors on $\beta_0, \beta_1, \dots, \beta_k$ and on $\sigma$. The posterior
$f(\beta_0,\dots,\beta_k,\sigma \mid y) \propto L(\beta_0,\dots,\sigma \mid y)\, f(\beta_0,\dots,\sigma)$
is now a *joint* distribution over all the coefficients at once. We summarize it the same way as before,
one coefficient at a time: report each coefficient's **posterior mean (or median) together with a
credible interval** — never a point alone. The interpretation of $\beta_j$ is the partial effect: the
expected change in $y$ per one-unit change in $x_j$, the other predictors held fixed.

## The logistic model statement

For a binary outcome the model is

$$
y_i \mid \pi_i \sim \text{Bernoulli}(\pi_i),
\qquad
\operatorname{logit}(\pi_i) = \log\!\frac{\pi_i}{1-\pi_i} = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik}.
$$

Inverting the link gives the success probability directly:

$$
\pi_i = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_{i1} + \cdots + \beta_k x_{ik})}}.
$$

Priors go on $\beta_0,\dots,\beta_k$ as usual, and the posterior
$f(\beta_0,\dots,\beta_k \mid y) \propto L(\beta_0,\dots,\beta_k \mid y)\, f(\beta_0,\dots,\beta_k)$ is a
joint distribution over the coefficients. There is no separate noise sd here: the Bernoulli variance
$\pi_i(1-\pi_i)$ is already pinned down by $\pi_i$.

## From a logistic coefficient to an odds ratio

The reason logistic coefficients are interpreted on the **odds** scale is a one-line piece of algebra,
and it is the most useful derivation of the week. Take a single predictor $x$ and compare the odds at
$x$ to the odds at $x+1$ (the other predictors, if any, held fixed). By the model,

$$
\operatorname{odds}(\pi \mid x) = \frac{\pi}{1-\pi} = e^{\,\beta_0 + \beta_1 x}.
$$

Step the predictor up by one unit:

$$
\operatorname{odds}(\pi \mid x+1) = e^{\,\beta_0 + \beta_1 (x+1)} = e^{\,\beta_0 + \beta_1 x}\, e^{\beta_1}.
$$

Divide the second by the first and the intercept and the $x$ term cancel:

$$
\frac{\operatorname{odds}(\pi \mid x+1)}{\operatorname{odds}(\pi \mid x)} = e^{\beta_1}.
$$

So $e^{\beta_1}$ is the **odds ratio** for a one-unit increase in $x$: a *multiplicative* factor on the
odds, not an additive change in the probability. A coefficient of $\beta_1 = 0$ gives odds ratio
$e^0 = 1$ (no effect); $\beta_1 > 0$ gives a ratio above $1$ (odds increase); $\beta_1 < 0$ gives a
ratio below $1$ (odds decrease). Because $\beta_1$ is a parameter with a posterior, the odds ratio
$e^{\beta_1}$ is itself a posterior quantity: transform each posterior draw of $\beta_1$ through
$\exp(\cdot)$ and you get a posterior — and hence a credible interval — for the odds ratio.

### Worked example — symbolic: a coefficient becomes an odds ratio

Take a one-predictor logistic model for whether a student passes a checkpoint ($y_i = 1$ for pass) as a
function of hours of focused review $x_i$:

$$
y_i \mid \pi_i \sim \text{Bernoulli}(\pi_i), \qquad \operatorname{logit}(\pi_i) = \beta_0 + \beta_1 x_i.
$$

Suppose the posterior for the slope is centered at $\beta_1 = 0.40$ with a $95\%$ credible interval
$[0.15,\ 0.66]$. On the odds scale the point summary is

$$
e^{\beta_1} = e^{0.40} \approx 1.49,
$$

and transforming the interval endpoints gives an odds-ratio credible interval

$$
\big[e^{0.15},\ e^{0.66}\big] \approx [1.16,\ 1.93].
$$

Reading: each additional hour of focused review multiplies the *odds* of passing by about $1.49$ — a
$49\%$ increase in the odds — and we are about $95\%$ posterior-confident the multiplier lies between
roughly $1.16$ and $1.93$. The whole interval sits above $1$, so the data, filtered through the prior,
push toward a positive association. Note what we did *not* say: we did not say the *probability* of
passing rises by a fixed amount per hour. It does not — the probability change depends on where you
start on the S-curve, which is exactly the point of the next example.

### Worked example — numeric instance you can re-run: a fitted logistic curve

The chunk below builds a small synthetic dataset (pass/fail versus review hours), fits a logistic model
with `glm`, draws the fitted S-curve, and prints the slope and its implied odds ratio. The `glm` fit is
a maximum-likelihood point summary; read the figure caption for how the **Bayesian** posterior-over-
curves view sits on top of it. All randomness is seeded, so you get the same numbers every time.

```{r}
#| label: fig-logistic-curve
#| fig-cap: "Synthetic pass/fail data (jittered points at y = 0 and y = 1) with the fitted logistic curve giving P(pass) as a function of review hours. The single solid curve is the maximum-likelihood (glm) fit; in the Bayesian view each posterior draw of the coefficients yields one such curve, so the honest picture is a *bundle* of curves whose vertical spread at any x is the posterior uncertainty in the predicted probability."
#| fig-alt: "Scatter of binary outcomes at heights zero and one across review hours from zero to ten, overlaid with a rising S-shaped logistic curve climbing from near zero probability on the left to near one on the right, crossing one-half around four to five hours."
set.seed(1010)
n <- 80
hours <- runif(n, 0, 10)                       # synthetic review hours
eta   <- -2.0 + 0.55 * hours                   # true log-odds (linear predictor)
pi    <- 1 / (1 + exp(-eta))                   # true success probabilities
y     <- rbinom(n, size = 1, prob = pi)        # observed pass/fail

fit <- glm(y ~ hours, family = binomial)       # logistic regression by ML
b   <- coef(fit)                               # intercept and slope estimates

# scatter with light vertical jitter so 0/1 points are visible
plot(hours, y + runif(n, -0.03, 0.03),
     pch = 19, col = rgb(0, 0, 0, 0.5),
     xlab = "Review hours (x)", ylab = "Pass (1) / Fail (0)  and  P(pass)",
     ylim = c(-0.05, 1.05), main = "Fitted logistic curve")

# overlay the fitted probability curve
xx <- seq(0, 10, length.out = 200)
pp <- 1 / (1 + exp(-(b[1] + b[2] * xx)))
lines(xx, pp, lwd = 2)
abline(h = 0.5, lty = 3); abline(v = -b[1] / b[2], lty = 3)  # where P(pass) = 0.5
legend("right", legend = "fitted P(pass)", lwd = 2, bty = "n")
```

```{r}
#| label: logistic-coefs
#| echo: true
slope    <- coef(fit)["hours"]
odds_rat <- exp(slope)                          # odds ratio per extra hour
round(c(slope = unname(slope), odds_ratio = unname(odds_rat)), 3)
```

The printed slope is the estimated log-odds change per extra hour; `exp(slope)` is the estimated odds
ratio. Because we *generated* the data with a true slope of $0.55$, the estimate should land near there,
and the odds ratio near $e^{0.55}\approx 1.73$. To make the Bayesian uncertainty concrete without extra
packages, you can also see the S-curve *steepest* near $P=0.5$ and *flat* at the extremes — which is why
a fixed change in $x$ moves the probability a lot in the middle and barely at all near $0$ or $1$.

## A transfer case — counts go to Poisson regression

The same link-the-linear-predictor recipe handles **count** outcomes. The clinic wait-time strand from
Week 7 is the natural carrier: suppose instead of wait times we count the **number of walk-ins per
hour**, $y_i \in \{0,1,2,\dots\}$, as a function of, say, the hour of day. A count cannot be Normal
(it is non-negative and discrete) and its mean must stay positive, so we use a **log link**:

$$
y_i \mid \lambda_i \sim \text{Poisson}(\lambda_i), \qquad \log(\lambda_i) = \beta_0 + \beta_1 x_{i1} + \cdots.
$$

Here $e^{\beta_1}$ is a **rate ratio**: a one-unit increase in $x_1$ multiplies the expected count by
$e^{\beta_1}$ — exactly parallel to the odds-ratio derivation above, with "odds" replaced by "rate." The
Bayesian machinery is identical: priors on the coefficients, a posterior over them, point-plus-interval
summaries, and prediction reported as a distribution. (Bayes Rules! Ch 12 develops this case in full;
we only flag it here so the pattern — pick a data model, link the linear predictor, keep the posterior —
is visible as one idea, not three.)

## A convention warning

Two cautions, both about reading coefficients honestly; treat them as the non-negotiable conventions of
the week.

The first convention concerns **log-odds versus probability**. A logistic coefficient $\beta_1$ lives on
the **log-odds** scale; $e^{\beta_1}$ is a multiplicative effect on the **odds**; neither is a change in
**probability**. The probability effect of a one-unit change in $x$ is *not constant* — it depends on
where you sit on the S-curve, large near $\pi = 0.5$ and tiny near $0$ or $1$. So "the coefficient is
$0.40$, so probability rises $0.40$" is simply wrong, and "the odds ratio is $1.49$, so probability
rises $49\%$" is also wrong (odds, not probability). Caution: state which scale you are on every single
time.

The second convention is the one this course repeats all term: **a prediction must carry its
uncertainty**, and you must say *which* uncertainty. A predicted probability or count from a Bayesian
model is a **posterior distribution**, summarized by a point estimate paired with a **credible interval**
— never a bare number, and never a *confidence* interval (a credible interval is a direct posterior
probability statement about the quantity; a confidence interval is a frequentist statement about the
procedure across hypothetical repetitions — do not conflate them). And distinguish the two *kinds* of
predictive uncertainty (SLO O11): uncertainty in the **fitted curve** (our posterior ignorance about the
coefficients, which shrinks as data grow) versus uncertainty in a **new individual outcome** (the
irreducible randomness of a single Bernoulli or Poisson draw, which does not vanish with more data). A
posterior-predictive interval for one new student folds in both; a credible interval for the success
probability folds in only the first.

## Practice (ungraded)

Use these to check your understanding. No answer keys appear here.

1. Write the one-predictor logistic model for whether a survey respondent bikes to campus as a function
   of distance-from-campus (km). Identify $\pi_i$, $\eta_i$, and which quantity is the parameter.
2. A logistic slope has posterior mean $-0.30$. Compute the implied odds ratio and translate it into one
   plain-English sentence about the odds. Is the association positive or negative?
3. Explain, in two sentences, why the *probability* change for a one-unit increase in a predictor is
   larger near $\pi = 0.5$ than near $\pi = 0.95$. A sketch of the S-curve may help.
4. For the multiple-predictor Normal model, state in words what "the posterior mean of $\beta_2$ is
   $1.7$ with credible interval $[0.4, 3.0]$" means, including the held-fixed clause.
5. A classmate reports a single predicted pass probability of $0.62$ with no interval. Name what is
   missing and which *two kinds* of uncertainty a full posterior-predictive statement would include.
6. Re-run the figure chunk with the true slope changed from $0.55$ to $0.20$. Predict, before running,
   whether the fitted S-curve gets steeper or flatter, then check.

## Formula-verification status

These formulas are prepared as evidence but NOT yet human/source verified (verified: false); see the
notation ledger. The course math gate is blocked pending sign-off. Specifically, the multiple-predictor
model statement, the logistic link and its inverse, the odds-ratio derivation ($e^{\beta_1}$), and the
Poisson log-link rate-ratio claim are staged as candidate results for review, not as confirmed course
truth. Do not treat any displayed identity here as final until the verification sign-off is recorded in
the notation ledger and the gate is cleared.

## Reading guide

- **Bayes Rules! Ch 11 (extending regression to several predictors)** supports the multiple-predictor section above.
- **Bayes Rules! Ch 13 (logistic regression)** supports the binary-outcome / log-odds section — read it to see the link function developed at length.
- **Bayes Rules! Ch 12 (Poisson regression)** supports the count-outcome transfer case we sketch. Use these chapters as enrichment for the models introduced here; the page already carries the core ideas.

## Public vs. graded

This page is a public, ungraded study note. Graded quiz prompts, their wording, point values, rubrics,
keys, and due dates are not here: the LMS (Blackboard) is authoritative for all graded specifics, and no
answer keys are posted here. The practice prompts above are for self-check only.

## Looking ahead

Next week — **Week 11: Model checking & comparison** — we stop assuming our regression is right and
start interrogating it: posterior predictive checks, comparing competing models, and deciding which one
to trust. See [Week 11 — Model checking & comparison](../notes/week-11-model-checking-comparison.qmd).

## See also

- [Lab 9 — Bayesian regression](../labs/lab-09-bayesian-regression.qmd) — the hands-on regression
  walkthrough whose workflow extends directly to the models above.
- [Notation glossary](../resources/notation-glossary.qmd) — the symbols and conventions this page inherits.
- [Bayes vs. classical cheatsheet](../resources/bayes-vs-classical-cheatsheet.qmd) — credible vs.
  confidence intervals, side by side.
- [Syllabus](../syllabus.qmd) — course outcomes (O9, O11) and the schedule.
