---
title: "Week 7 — Simulation-first computation"
subtitle: "Grid approximation, posterior draws, and a reproducible Quarto workflow"
page_mode: software_lab_week
week_contract: software_note_lab
source_mode: source-informed
readings:
  - "Bayes Rules! Ch 6 — Approximating the posterior (grid & MCMC) — supports grid approximation, posterior draws, and the idea that MCMC is a tool for sampling a posterior we cannot write down."
toc: true
---

## The idea

Up to now we have leaned on **conjugacy**: a Beta prior met a Binomial likelihood and the posterior
came out as another Beta, with parameters we could update by hand. That is a gift, not a guarantee.
Most realistic models are not conjugate — the posterior $f(\theta \mid y)$ exists, but no clean
formula falls out. The central concept of this week is that we do not actually *need* a formula. If we
can **evaluate** the unnormalized posterior at many values of $\theta$, or **sample** from the
posterior, we can answer every question we care about — posterior mean, credible interval, predictive
checks — by computing on those numbers instead of solving an integral.

That reframing is the whole idea: **a posterior is something you can compute with, not just something
you write down.** We will meet two computational paths. **Grid approximation** chops the parameter
space into a fine grid, evaluates prior $\times$ likelihood at each grid point, normalizes, and then
treats the result as a discrete approximation we can sample from. **Markov chain Monte Carlo (MCMC)**
takes a cleverer route that scales to many parameters, and we introduce it conceptually — what it does
and how to ask "did it converge?" — without the theory. Both produce the same kind of output we will
reuse all term: a big pile of **posterior draws** of $\theta$.

Because the right answer for our recurring bike-proportion problem is *already known* (it is conjugate,
so we have the exact Beta), this week is the perfect place to learn the machinery: we can check the
approximation against the truth. That is the discipline of simulation-first computation — always
compute the thing you can verify before you trust the thing you cannot.

## Where we are and why this matters

In Weeks 3–4 we built the **Beta-Binomial** model for an unknown proportion $p$: the proportion of
students who bike to campus. With a mild prior $\text{Beta}(2,2)$ and a synthetic survey of
**8 bikers out of 24** surveyed, the conjugate update gave a posterior $\text{Beta}(10,18)$ with
posterior mean $10/28 \approx 0.357$. In Weeks 5–6 we stress-tested that posterior with
**prior sensitivity** and built **posterior-predictive** reasoning — asking what *future* survey
counts the model expects.

This week takes those exact same ideas and **computes** them rather than solving them. We will
grid-approximate the $\text{Beta}(10,18)$ posterior and confirm the grid recovers the mean we already
know. We will draw thousands of samples and read the posterior off the histogram. The payoff is
general: once you can approximate a posterior you can write down, you can approximate one you cannot —
which is exactly what real models demand. This is **the computation week** of the course, and the
habits you build here (seeds, code-with-interpretation, rendered reports) carry into every later week,
especially regression (Weeks 9–11) and hierarchical models (Week 13), where conjugacy is gone.

## Learning goals

By the end of this week you should be able to:

- Explain why we approximate a posterior at all, and when conjugacy is *not* available.
- Build a **grid approximation**: evaluate prior $\times$ likelihood on a grid, normalize, and sample.
- Read posterior summaries (mean, credible interval) off a **set of posterior draws**.
- Describe, in plain language, what **MCMC** does and what "did it converge?" is checking.
- Assemble a **reproducible Quarto report**: seeded code, output, and interpretation together, with
  session info, rendered to HTML.
- Recognize and fix the common runtime failures (R not found, unseeded simulation, a chunk error).

These goals serve course outcomes **O5** (compute and summarize a posterior) and **O13**
(produce a reproducible computational workflow).

## Setup

**First,** make sure your local toolchain is alive: R computes, VS Code edits, and Quarto renders.
If you have not done this, work through the [setup page](../resources/r-vscode-quarto-setup.qmd)
before going further — a render that cannot find R will stop you cold, and we would rather discover
that now than during the lab.

**Then,** open a fresh `.qmd` and confirm a trivial chunk runs. Everything this week uses **base R
only** — no add-on packages — so if R is discoverable, you have everything you need. We will rely on
`seq`, `dbeta`, `sample`, `mean`, `quantile`, `hist`, `curve`, and `lines`. Keep a habit from the very
first chunk: **set a seed**. Any line of code that draws random numbers must be preceded (in the same
reproducible document) by `set.seed(...)`, so that "run it again" gives the same figure.

## Grid approximation, step by step

A grid approximation turns a continuous posterior into a fine **discrete** one. The recipe is short and
every step is something base R does in one line.

1. **Lay down a grid.** Choose many evenly spaced candidate values of $\theta$ across its range. For a
   proportion, $\theta \in [0,1]$, so a grid like `seq(0, 1, length.out = 1001)` works.
2. **Evaluate prior $\times$ likelihood at each grid point.** The unnormalized posterior is
   $L(\theta \mid y)\, f(\theta)$. For the bike problem, $f(\theta)$ is the $\text{Beta}(2,2)$ density
   and $L(\theta \mid y)$ is the Binomial likelihood for 8 successes in 24 trials. Multiply them.
3. **Normalize.** Divide each value by the sum of all of them, so the grid values add to 1. This is the
   discrete stand-in for dividing by the evidence $f(y)$ — the constant that $\propto$ usually drops in
   $f(\theta \mid y) \propto L(\theta \mid y)\, f(\theta)$. On a grid we get to compute it.
4. **Sample.** Draw grid values *with replacement*, with probabilities equal to the normalized weights.
   The result is a set of posterior draws you can summarize like any data: `mean`, `quantile`, `hist`.

The finer the grid and the more draws, the closer this gets to the exact posterior. The beauty is that
nothing in steps 1–4 used the fact that the bike model is conjugate. The same four steps work for a
posterior with no closed form — you just swap in whatever prior and likelihood you have.

```{r}
#| label: fig-grid-vs-exact
#| fig-cap: "Grid-approximated posterior for the bike proportion (points) against the exact Beta(10,18) posterior (curve). The grid recovers the closed-form shape; both peak near 0.36."
#| fig-alt: "A density curve for the exact Beta(10,18) posterior rising from near zero, peaking just below 0.36, and falling toward 0.7, with small open circles from the grid approximation sitting almost exactly on the curve. A vertical dashed line marks the posterior mean at about 0.357."
# Bike proportion: prior Beta(2,2), data 8 successes in 24 trials -> exact Beta(10,18)
theta <- seq(0, 1, length.out = 1001)        # the grid
prior <- dbeta(theta, 2, 2)                    # Beta(2,2) prior density
like  <- dbinom(8, size = 24, prob = theta)    # Binomial likelihood, y=8, n=24
unnorm <- like * prior                          # prior x likelihood (unnormalized posterior)
post_grid <- unnorm / sum(unnorm)               # normalize so weights sum to 1

# Plot the EXACT posterior as a density, overlay the grid as a comparable density
exact <- dbeta(theta, 10, 18)
plot(theta, exact, type = "l", lwd = 2,
     xlab = expression(theta ~ "(proportion who bike)"), ylab = "posterior density",
     main = "Grid approximation vs exact Beta(10,18)")
# scale grid weights back to density units (divide by grid spacing) for an honest overlay
grid_density <- post_grid / (theta[2] - theta[1])
points(theta[seq(1, 1001, by = 20)], grid_density[seq(1, 1001, by = 20)],
       pch = 1, cex = 0.7)
abline(v = 10/28, lty = 2)
legend("topright", legend = c("exact Beta(10,18)", "grid approximation", "posterior mean"),
       lty = c(1, NA, 2), pch = c(NA, 1, NA), lwd = c(2, NA, 1), bty = "n")
```

The grid points sit essentially on top of the exact curve. That visual agreement is the verification:
when the truth is known, the approximation reproduces it, which licenses us to trust the method when the
truth is *not* known.

## Posterior draws are the working currency

Once we have sampled, we no longer talk to a formula — we talk to a **vector of draws**. Every summary
is now an ordinary computation on that vector. The posterior mean is `mean(draws)`. A 95% **credible
interval** is `quantile(draws, c(0.025, 0.975))` — and notice we say *credible* interval, the
probability statement about $\theta$ given the data, never *confidence* interval. The posterior median
is `median(draws)`. We always report a point estimate **with** an interval, never a point alone.

```{r}
#| label: fig-grid-draws
#| fig-cap: "Histogram of 20,000 posterior draws sampled from the grid approximation of the bike-proportion posterior, with the simulated mean (solid) and a 95 percent credible interval (dashed). The simulated mean lands at about 0.357, matching the closed-form value."
#| fig-alt: "A roughly bell-shaped histogram of posterior draws centered near 0.36, ranging from about 0.2 to 0.55. A solid vertical line marks the simulated mean near 0.357; two dashed vertical lines mark the lower and upper bounds of the 95 percent credible interval at roughly 0.20 and 0.53."
set.seed(2026)
theta <- seq(0, 1, length.out = 1001)
prior <- dbeta(theta, 2, 2)
like  <- dbinom(8, size = 24, prob = theta)
post_grid <- (like * prior) / sum(like * prior)

draws <- sample(theta, size = 20000, replace = TRUE, prob = post_grid)
sim_mean <- mean(draws)
ci <- quantile(draws, c(0.025, 0.975))

hist(draws, breaks = 40, freq = FALSE, col = "grey90", border = "white",
     xlab = expression(theta), main = "20,000 posterior draws (grid-sampled)")
abline(v = sim_mean, lwd = 2)
abline(v = ci, lty = 2)
legend("topright",
       legend = c(sprintf("simulated mean = %.3f", sim_mean),
                  sprintf("95%% CI = [%.3f, %.3f]", ci[1], ci[2])),
       lty = c(1, 2), lwd = c(2, 1), bty = "n")
```

The simulated mean prints at about **0.357** — the same $10/28$ we computed by hand in Week 3. That
match is not luck; it is the point. When you cannot check against a formula, this same code still
produces a defensible mean and interval, and you trust it *because* you watched it pass the check here.

## MCMC, conceptually

Grid approximation is wonderful for one or two parameters and miserable for many: a grid over ten
parameters has astronomically many cells. **MCMC** is the workhorse that scales. The concept, stripped
of theory: instead of evaluating the posterior everywhere, MCMC takes a guided **random walk** through
the parameter space that is designed to **spend time in each region in proportion to its posterior
probability**. The sequence of places it visits — the **chain** — becomes your set of posterior draws,
and you summarize it exactly as before: `mean`, `quantile`, `hist`.

You do not need the algorithm to use it responsibly, but you do need one habit: **ask "did it
converge?"** A chain has to "forget" where it started and settle into sampling the real posterior.
Two cheap diagnostics, which the course packages compute for you:

- **Trace plots** — a plot of the chain's value over its iterations. A converged chain looks like a
  fuzzy, stationary horizontal band ("a hairy caterpillar"); a chain that is wandering, trending, or
  stuck has not converged.
- **Running several chains from different starts** and checking they overlap. If independent chains land
  on the same distribution, that is evidence they found it. (The numeric summary of this is called
  R-hat; you want it near 1.)

We introduce MCMC now so the *idea* is in place; the dedicated tooling appears later in the course. The
mental model to keep: **MCMC gives you posterior draws, and draws are something you already know how to
summarize.** Below is illustrative code in the style the course packages use — it is shown, not run.

```r
# illustrative — runs with the course packages (rstanarm), not executed here
library(rstanarm)
fit <- stan_glm(bikes ~ 1, data = survey, family = binomial,
                prior_intercept = normal(0, 1), seed = 2026,
                chains = 4, iter = 2000)
plot(fit, "trace")        # eyeball convergence: hairy caterpillars
summary(fit)              # look for Rhat near 1
posterior <- as.data.frame(fit)   # the draws — summarize like any vector
```

## Reproducible workflow

A result you cannot regenerate is a rumor. The deliverable of this week's habits is a single **`.qmd`
file** that holds code, output, and interpretation together and **renders** to a clean report. The
organizing principles:

- **One file, one story.** The `.qmd` is the source of truth. Prose, code chunks, and the figures they
  produce live in the same document, in reading order. A reader (and future you) sees the claim and the
  code that backs it side by side.
- **Seed every simulation.** Put `set.seed(...)` in any chunk that draws random numbers, so that
  `render` reproduces the exact figure. An unseeded simulation gives a different answer each render —
  fine for nothing you want to defend.
- **Interpret next to the number.** After each figure or summary, write one or two sentences saying what
  it means. Code that is never interpreted is not yet a result.
- **Save a session record.** End the report with a chunk that prints `sessionInfo()` (R version and
  loaded packages). When a future render breaks, the session record is your first clue.
- **Render to share.** Run `quarto render report.qmd` to produce the HTML (or PDF). You hand in the
  rendered report *and* the `.qmd` so a grader can re-run it.

```{r}
#| label: session-record
#| echo: true
# A session record at the foot of every reproducible report
sessionInfo()
```

That single rendered `.qmd` — seeded chunks, figures, interpretation, session info — is the artifact the
rest of the course expects you to be able to produce on demand.

## When it breaks

Computation fails in a few predictable ways. Recognizing the symptom is most of the fix.

- **"Quarto can't find R" / the render errors before any chunk runs.** Quarto must locate R to execute
  `{r}` chunks. If the render fails complaining it cannot find R, R is **off your PATH**. Fix: put R's
  `bin` on PATH, set the `QUARTO_R` environment variable, or render from a shell whose PATH already
  includes R's `bin`. (This is exactly the runner's situation in this course: R 4.6.0 is installed but
  off PATH, so the build prepends R's bin before `quarto render`.) See the
  [setup page](../resources/r-vscode-quarto-setup.qmd).
- **A figure changes every time you render — an unseeded simulation.** Symptom: your mean and interval
  drift between renders. Troubleshoot by searching for any `sample`, `rbeta`, `rnorm` not preceded by
  `set.seed(...)` in the same chunk or document. Add the seed; the figure freezes.
- **A chunk error halts the whole render.** One broken chunk stops the document — nothing after it
  builds. Fix by isolating: run the failing chunk interactively in the R console, read the error
  message (a misspelled object, a wrong argument name, a package that is not installed), correct it, then
  re-render. Keep chunks small so the failing line is easy to find.
- **"Object not found."** Usually a chunk depends on a variable defined in a chunk that did not run (or
  ran in a different order). Define what you use within the document's top-to-bottom flow.

The throughline: if a figure is missing or wrong, ask first *was R found?* and *was it seeded?* before
suspecting the statistics.

## A hands-on step

Reading about grids is not the same as building one. This week's **hands-on** practice is
[Lab 7 — Grid approximation in Quarto](../labs/lab-07-grid-approximation-quarto.qmd), where you
approximate the bike-proportion posterior on a grid, draw from it, compare the simulated mean to the
closed-form $10/28$, and assemble the whole thing as a reproducible report. Do the lab in a real `.qmd`
and render it — the friction you hit while rendering is the lesson.

## AI use note

Generative AI can speed up boilerplate, but in a reproducible workflow you remain accountable for every
number. Use it transparently:

- **Tool.** A general coding assistant (e.g. an LLM in your editor).
- **Purpose.** Allowed: scaffolding a chunk, explaining an R error message, drafting a caption,
  suggesting a base-R idiom. Not a substitute for understanding what the grid or the draws mean.
- **Verification.** You must verify every AI-suggested line yourself: run it, confirm the simulated mean
  matches the closed-form posterior mean ($10/28 \approx 0.357$) where a check exists, and confirm the
  chunk is seeded. Disclose AI assistance as your course's AI policy (in the LMS) requires. AI output
  that you cannot verify against a known quantity does not belong in a report you sign.

## Portfolio connection

The rendered report you produce this week is a natural **portfolio** piece. It demonstrates outcome
**O13** directly: a seeded, interpreted, reproducible Quarto document that someone else could re-run and
get your numbers. Keep the `.qmd` and its rendered HTML; later portfolio entries (regression, hierarchical
models) build on exactly this skeleton, so a clean Week 7 report pays off all term.

## Reading guide

- **Bayes Rules! Ch 6 (Approximating the posterior — grid & MCMC)** maps onto this week section by section: its grid treatment supports our grid-approximation walkthrough, and its MCMC introduction supports the *MCMC, conceptually* discussion (what sampling buys you when conjugacy is unavailable, and the "did it converge?" habit). Read it alongside Lab 7 — it is enrichment, not a prerequisite; the page and lab stand on their own.

## Public vs. graded

This page and Lab 7 are **public study materials and ungraded practice**. The graded version of this
week's work — the deliverable, its rubric, point values, and due date — lives in the LMS. For anything
that affects your grade, **the LMS (Blackboard) is authoritative**, and **no answer keys are posted
here**. If a public page and the LMS ever disagree, follow the LMS.

## Looking ahead

Next week is **Week 8 — Midterm synthesis**: we pause new machinery and pull the first half of the
course together — prior, likelihood, posterior, prediction, and now computation — into one coherent
picture before the midterm. Bring your Week 7 report; the habits you built here are part of what we
synthesize.

## See also

- [Lab 7 — Grid approximation in Quarto](../labs/lab-07-grid-approximation-quarto.qmd)
- [R · VS Code · Quarto setup](../resources/r-vscode-quarto-setup.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)
- [Week 8 — Midterm synthesis](../notes/week-08-midterm-synthesis.qmd)
- [Syllabus](../syllabus.qmd)
