---
title: "Week 4 — The Beta-Binomial model"
subtitle: "Why a Beta prior and Binomial data give a Beta posterior, and how to read it"
page_mode: math_derivation_week
week_contract: math_derivation
source_mode: source-informed
readings:
  - "Bayes Rules! Ch 3 — The Beta-Binomial model — supports the conjugate update and the Beta posterior"
  - "Bayes Rules! Ch 4 — Balance and sequentiality — supports how prior strength and sample size trade off, and sequential updating"
toc: true
---

## Mathematical goal

This week we make precise a claim that Week 3 only sketched: when we model an unknown
proportion $\theta$ with a Beta prior and observe Binomial data, the posterior is again a
Beta distribution, and we can write down its parameters without doing a single integral. The
goal is to **derive** that result by recognizing the *kernel* of a Beta density inside the
product $L(\theta \mid y)\, f(\theta)$, and then to read the resulting posterior responsibly —
pairing a point summary with a credible interval.

Concretely, by the end you should be able to start from a Beta($\alpha,\beta$) prior, observe
$y$ successes in $n$ trials, and produce the posterior Beta($\alpha+y,\ \beta+n-y$) together
with its mean $(\alpha+y)/(\alpha+\beta+n)$ — and explain *why* each piece appears.

## The week question

If we believe a proportion $\theta$ lives somewhere on $[0,1]$ and then we watch $y$ successes
out of $n$ tries, exactly which Beta distribution describes what we now believe about $\theta$ —
and how do we summarize it honestly?

## Where we are and why this matters

In Week 3 we built the machinery of **prior, likelihood, posterior** on the recurring
bike-to-campus survey: we want $\theta$, the proportion of students who bike to campus, we
encoded a starting belief as a prior $f(\theta)$, and we let data reshape it through the central
identity

$$
f(\theta \mid y) \;\propto\; L(\theta \mid y)\, f(\theta).
$$

That identity is true in general, but in general the proportionality constant — the evidence
$f(y)$ — requires an integral that we cannot always do by hand. This week we cash in a piece of
good luck: for proportions, a Beta prior and a Binomial likelihood fit together so cleanly that
the integral takes care of itself. This pairing is called **conjugacy**, and it is the first
fully worked, closed-form Bayesian model in the course. Everything later (prior sensitivity in
Week 5, prediction in Week 6, simulation in Week 7) builds on being able to write this posterior
down exactly.

This matters because it turns Bayesian updating from an abstraction into arithmetic: "two new
numbers" instead of "an intractable integral."

## Notation

This week leans on the fixed course notation. The table below is the working ledger; the words
in it are the ones the derivation manipulates.

| Symbol | Meaning |
|---|---|
| $\theta$ | the unknown parameter — here a proportion in $[0,1]$ |
| $y$ | observed data — the number of successes |
| $n$ | the number of trials (fixed in advance) |
| $f(\theta)$ | the prior pdf on $\theta$ |
| $L(\theta \mid y)$ | the likelihood — a **function of $\theta$**, not a distribution over $\theta$ |
| $f(\theta \mid y)$ | the posterior pdf on $\theta$ |
| $f(y)$ | the evidence / marginal — the normalizing constant $\propto$ drops |
| Beta($\alpha,\beta$) | Beta with **shape-shape** parameters $\alpha,\beta>0$ |
| $\propto$ | "proportional to": equal up to a factor not depending on $\theta$ |

A reminder on parameterization that will matter below: we always write Beta($\alpha,\beta$) in
the **shape-shape** convention. Both $\alpha$ and $\beta$ are shapes; neither is a rate.

## Conceptual setup

Before any algebra, recall the two ingredients and what we assume about each.

**The prior.** We assume $\theta \sim$ Beta($\alpha,\beta$). The Beta family is supported on
$[0,1]$, which is exactly the home of a proportion, so it is a natural language for prior belief
about $\theta$. Its density is

$$
f(\theta) \;=\; \frac{1}{B(\alpha,\beta)}\,\theta^{\alpha-1}(1-\theta)^{\beta-1},
\qquad 0 \le \theta \le 1,
$$

where $B(\alpha,\beta)$ is the Beta function, a constant that makes the density integrate to 1.
The part that depends on $\theta$ — the **kernel** — is $\theta^{\alpha-1}(1-\theta)^{\beta-1}$.
Hold onto that shape.

**The likelihood.** We assume the data model is Binomial: given $\theta$, the count of
successes $Y$ in $n$ independent trials each with success probability $\theta$ has

$$
L(\theta \mid y) \;=\; \binom{n}{y}\,\theta^{y}(1-\theta)^{n-y}.
$$

Read as a function of $\theta$ (with $y$ and $n$ fixed by the data), the $\binom{n}{y}$ is just a
constant. The kernel in $\theta$ is $\theta^{y}(1-\theta)^{n-y}$.

The setup is now suggestive: both kernels are powers of $\theta$ times powers of $(1-\theta)$.
When we multiply them, the exponents simply add. That is the whole trick.

## The derivation by kernel recognition

Start from the central identity and substitute the two pieces, keeping only what depends on
$\theta$ (every constant gets absorbed into $\propto$):

$$
f(\theta \mid y)
\;\propto\; L(\theta \mid y)\, f(\theta)
\;\propto\; \theta^{y}(1-\theta)^{n-y}\cdot \theta^{\alpha-1}(1-\theta)^{\beta-1}.
$$

Here $\propto$ has dropped two things: the binomial coefficient $\binom{n}{y}$ from the
likelihood, and the Beta-function constant $1/B(\alpha,\beta)$ from the prior — neither depends
on $\theta$, and the evidence $f(y)$ (also a constant in $\theta$) is what would restore equality.
Now combine the like bases by adding exponents:

$$
f(\theta \mid y)
\;\propto\; \theta^{(\alpha+y)-1}\,(1-\theta)^{(\beta+n-y)-1}.
$$

Compare this to the Beta kernel $\theta^{a-1}(1-\theta)^{b-1}$. It matches **exactly**, with

$$
a = \alpha + y, \qquad b = \beta + n - y.
$$

A density is determined by its kernel (the normalizing constant is forced once the kernel is
fixed), so we can read off the answer without integrating:

$$
\boxed{\,\theta \mid y \;\sim\; \text{Beta}(\alpha + y,\ \beta + n - y).\,}
$$

This is **conjugacy**: a Beta prior plus Binomial data yields a Beta posterior. The update rule
is almost embarrassingly simple — add your successes $y$ to $\alpha$, add your failures $n-y$ to
$\beta$.

### The posterior mean

The mean of a Beta($a,b$) is $a/(a+b)$. Plugging in our posterior shapes:

$$
\mathbb{E}[\theta \mid y]
= \frac{\alpha + y}{(\alpha + y) + (\beta + n - y)}
= \frac{\alpha + y}{\alpha + \beta + n}.
$$

It is worth seeing this as a **weighted average** of the prior mean $\alpha/(\alpha+\beta)$ and
the data proportion $y/n$:

$$
\frac{\alpha + y}{\alpha + \beta + n}
= \underbrace{\frac{\alpha+\beta}{\alpha+\beta+n}}_{\text{weight on prior}}\cdot\frac{\alpha}{\alpha+\beta}
\;+\;
\underbrace{\frac{n}{\alpha+\beta+n}}_{\text{weight on data}}\cdot\frac{y}{n}.
$$

As $n$ grows, the weight on the data goes to 1 and the prior fades — the *balance* idea from
Bayes Rules! Chapter 4. The quantity $\alpha+\beta$ behaves like a "prior sample size": a large
$\alpha+\beta$ is a confident prior that the data must work harder to move.

## Credible intervals from the Beta posterior

A point estimate alone hides our uncertainty, so we always pair the posterior mean (or median)
with a **credible interval**. A 95% credible interval is an interval $[\ell, u]$ such that

$$
P(\ell \le \theta \le u \mid y) = 0.95.
$$

For a Beta posterior the equal-tailed version is immediate from quantiles: take the 2.5th and
97.5th percentiles of Beta($\alpha+y,\ \beta+n-y$). In R these are `qbeta(0.025, a, b)` and
`qbeta(0.975, a, b)`. We can also approximate the same interval by simulation — draw many values
with `rbeta` and read off the empirical quantiles — which is the bridge to Week 7.

The interpretation is direct and is one of the genuine payoffs of the Bayesian framing: given
the data and the model, there is a 95% posterior probability that $\theta$ lies in the interval.
That is a statement about $\theta$, which is what we usually wanted all along.

## Worked examples

We work two cases: the recurring bike survey (symbolic, then numeric) and a transfer case in a
different context, to show the update is about structure, not the story.

### Worked example — bike-to-campus proportion (symbolic, then numeric)

This is the recurring case. We model $\theta$, the proportion of students who bike to campus.

**Symbolic.** Take a mild prior Beta($\alpha,\beta$) = Beta(2, 2), which is symmetric and gently
favors values near $1/2$ without much conviction (prior sample size $\alpha+\beta = 4$). Suppose
the survey yields $y$ bikers out of $n$ respondents. By the derived rule,

$$
\theta \mid y \;\sim\; \text{Beta}(2 + y,\ 2 + n - y),
\qquad
\mathbb{E}[\theta \mid y] = \frac{2 + y}{4 + n}.
$$

**Numeric instance.** Now put in the recurring numbers: $y = 8$ bikers out of $n = 24$
respondents (so $n - y = 16$ non-bikers). The update adds successes to $\alpha$ and failures to
$\beta$:

$$
\theta \mid y \;\sim\; \text{Beta}(2 + 8,\ 2 + 16) = \text{Beta}(10,\ 18).
$$

Posterior mean:

$$
\mathbb{E}[\theta \mid y] = \frac{10}{10 + 18} = \frac{10}{28} \approx 0.357.
$$

Compare the inputs: the prior mean was $2/4 = 0.5$ and the raw data proportion was
$8/24 \approx 0.333$. The posterior mean $0.357$ sits between them, pulled most of the way toward
the data because $n = 24$ outweighs the prior sample size of 4. The figure below shows the prior
and the posterior together; the second figure shows the simulated posterior with its 95% credible
interval shaded.

```{r}
#| label: fig-prior-posterior-beta
#| fig-cap: "Prior Beta(2,2) and posterior Beta(10,18) for the bike-to-campus proportion. The posterior is tighter and centered near 0.357; the prior is broad and centered at 0.5."
#| fig-alt: "Two density curves on the interval 0 to 1. A low, broad dashed curve labeled prior Beta(2,2) peaks gently around 0.5. A taller, narrower solid curve labeled posterior Beta(10,18) peaks around 0.36, well to the left of the prior peak, showing the data sharpened belief and pulled it toward the data proportion."
theta <- seq(0, 1, length.out = 400)
prior <- dbeta(theta, 2, 2)
post  <- dbeta(theta, 10, 18)
plot(theta, post, type = "l", lwd = 2, lty = 1,
     xlab = expression(theta ~ "(proportion who bike)"),
     ylab = "density",
     main = "Prior vs posterior (bike-to-campus)")
lines(theta, prior, lwd = 2, lty = 2)
abline(v = 10/28, lwd = 1, lty = 3)
legend("topright",
       legend = c("posterior Beta(10,18)", "prior Beta(2,2)", "post. mean 0.357"),
       lty = c(1, 2, 3), lwd = c(2, 2, 1), bty = "n")
```

```{r}
#| label: fig-posterior-sim-ci
#| fig-cap: "Simulated posterior draws for the bike proportion (50,000 rbeta draws from Beta(10,18)) as a histogram, with the equal-tailed 95% credible interval shaded and its endpoints from qbeta marked."
#| fig-alt: "A histogram of simulated posterior values of the bike proportion, mound-shaped and centered near 0.36. A shaded region in the middle marks the central 95 percent credible interval running from roughly 0.20 to 0.53, with vertical lines at the two endpoints and a vertical line at the posterior mean near 0.357."
set.seed(404)
draws <- rbeta(50000, 10, 18)
ci <- qbeta(c(0.025, 0.975), 10, 18)
h <- hist(draws, breaks = 40, plot = FALSE)
plot(h, freq = FALSE, col = "white", border = "grey60",
     xlab = expression(theta ~ "(simulated posterior draws)"),
     ylab = "density",
     main = "Simulated posterior with 95% credible interval")
in_ci <- h$mids >= ci[1] & h$mids <= ci[2]
rect(h$breaks[-length(h$breaks)][in_ci], 0,
     h$breaks[-1][in_ci], h$density[in_ci],
     col = "grey80", border = "grey60")
abline(v = ci, lwd = 2, lty = 2)
abline(v = mean(draws), lwd = 2, lty = 1)
legend("topright",
       legend = c(sprintf("95%% CI [%.3f, %.3f]", ci[1], ci[2]),
                  sprintf("sim. mean %.3f", mean(draws))),
       lty = c(2, 1), lwd = 2, bty = "n")
```

The simulated mean lands within rounding of the closed-form $10/28 \approx 0.357$, and the
shaded interval matches `qbeta(c(0.025, 0.975), 10, 18)` — a sanity check we will lean on in
the lab.

### Worked example — transfer case (sprouted seeds, a different proportion)

To show the update is about the *structure*, not the bikes, transfer it to a synthetic
germination question: $\theta$ is the proportion of seeds from a batch that sprout. We have no
strong prior, so use a flat Beta(1, 1) (the uniform prior — every proportion equally plausible).
We plant $n = 40$ seeds and observe $y = 30$ sprouts ($n - y = 10$ duds).

Apply the same rule — successes onto $\alpha$, failures onto $\beta$:

$$
\theta \mid y \;\sim\; \text{Beta}(1 + 30,\ 1 + 10) = \text{Beta}(31,\ 11),
\qquad
\mathbb{E}[\theta \mid y] = \frac{31}{42} \approx 0.738.
$$

With a flat prior the posterior mean $31/42 \approx 0.738$ sits very close to the raw proportion
$30/40 = 0.75$; the tiny shift toward $1/2$ is the residual pull of the Beta(1,1) "one prior
success, one prior failure." An equal-tailed 95% credible interval is
`qbeta(c(0.025, 0.975), 31, 11)`, roughly $[0.59, 0.86]$. We would report this as: "the
posterior mean germination rate is about 0.74, 95% credible interval about $[0.59, 0.86]$" —
mean **and** interval, never the mean alone.

## A convention warning

Three conventions cause most of the avoidable errors here; treat them as cautions.

- **Shape-shape, not shape-rate.** Beta($\alpha,\beta$) uses *two shapes*. Do not import the
  Gamma habit of treating the second argument as a rate. In R, `dbeta(x, shape1, shape2)`,
  `qbeta`, and `rbeta` all expect the two Beta shapes in that order. Swapping them mirrors the
  distribution and silently corrupts every summary.

- **Credible interval is not a confidence interval.** A 95% **credible** interval is a posterior
  probability statement about $\theta$: given the data, $P(\ell \le \theta \le u \mid y) = 0.95$.
  A frequentist 95% **confidence** interval is a statement about a procedure's long-run coverage
  over hypothetical repeated samples, *not* a probability that this particular interval contains
  $\theta$. They can even look numerically similar yet mean different things — never conflate them,
  and always say "credible interval" for a posterior.

- **What $\propto$ drops.** In the derivation, $\propto$ discarded $\binom{n}{y}$ and
  $1/B(\alpha,\beta)$ and stood in for dividing by the evidence $f(y)$. That is legitimate only
  because none of those depend on $\theta$. If you ever drop something that *does* depend on
  $\theta$, the kernel is wrong and so is the posterior.

## Practice (ungraded)

Use these to check your understanding; no answer keys are posted here.

1. Starting from a Beta(3, 5) prior, you observe 12 successes in 20 trials. Write the posterior
   Beta parameters and its mean. Is the mean closer to the prior mean or the data proportion, and
   why?
2. In the bike example, suppose a second wave of the survey adds 5 more bikers out of 15 more
   respondents. Update the Beta(10, 18) posterior again (sequential updating). What single
   Beta would you have gotten by pooling all responses at once?
3. Explain, in one sentence each, what the $\propto$ step is allowed to drop and what it is *not*
   allowed to drop.
4. Sketch (by hand or with `curve(dbeta(x, 31, 11), 0, 1)`) the germination posterior and mark
   roughly where its 95% credible interval falls.
5. A classmate reports "the posterior mean is 0.357" with no interval. Write the one extra thing
   they must report and why a point estimate alone is not enough.

## Formula-verification status

These formulas are prepared as evidence but NOT yet human/source verified
(verified: false); see the notation ledger. The course math gate is blocked pending sign-off.
In particular, the conjugate update Beta($\alpha+y,\ \beta+n-y$), the posterior mean
$(\alpha+y)/(\alpha+\beta+n)$, and the credible-interval quantiles are staged here as derivation
evidence; they should be treated as provisional until the math gate is cleared and a reviewer's
sign-off is recorded against the ledger.

## Public vs. graded

This is a public, ungraded study note: no answer keys are posted here. Worked examples, practice
prompts, and figures are for learning, not for credit. Anything graded — quiz items, their
prompts and keys, weights, rubrics, point values, and due dates — lives in the LMS.
The LMS (Blackboard) is authoritative for all of it. If this page and the LMS ever disagree, follow the LMS.

## Looking ahead

Next week (Week 5 — Prior sensitivity & summaries) we hold the data fixed and ask how much the
choice of Beta prior actually changes the posterior, and we sharpen how we summarize a posterior
beyond the mean. See [Week 5 — Prior sensitivity & summaries](../notes/week-05-prior-sensitivity-summaries.qmd).

## See also

- [Lab 4 — Beta-Binomial simulation](../labs/lab-04-beta-binomial-simulation.qmd)
- [Notation glossary](../resources/notation-glossary.qmd)
- [Bayes vs. classical cheatsheet](../resources/bayes-vs-classical-cheatsheet.qmd)
- [Syllabus](../syllabus.qmd)
