---
title: "Week 2 — Discrete Bayes & diagnostic reasoning"
subtitle: "Why a positive test can still mean you probably don't have the disease"
page_mode: normal_week
week_contract: public_note
source_mode: source-informed
readings:
  - "Bayes Rules! Ch 2 — Bayes' Rule — supports conditional probability, the prior/likelihood/marginal pieces of Bayes' rule, and the diagnostic-test intuition we develop this week."
toc: true
---

## The week question

If you take a test for a rare condition and it comes back **positive**, what is the probability you
actually have the condition?

The honest answer surprises almost everyone: it can be *low* — sometimes well below 50% — even when
the test is "95% accurate." This week we build the machinery that explains why, and in doing so we
meet **Bayes' rule** in its cleanest, fully discrete form. Everything later in the course is a richer
version of the move we make here.

## Where we are and why this matters

In Week 1 we framed Bayesian thinking as **plausibility updating**: you start with what you believe
about an unknown, you observe evidence, and you revise your belief so it accounts for that evidence.
That was the *idea*. This week we make the idea **mechanical and checkable** in the simplest setting,
where the unknown takes only a few discrete values and we can write every probability in a small
table.

Discrete Bayes is the right place to start for two reasons. First, the arithmetic is fully visible —
nothing is hidden inside an integral, so you can audit every number. Second, the diagnostic-test
example forces a confrontation with **base rates** (how common the condition is *before* testing),
which is exactly the ingredient people drop when they reason badly about evidence. Get the discrete
case right and the continuous case in Week 3 is the same story with a smooth curve in place of a
table.

## Learning goals

By the end of this week you should be able to:

- State conditional probability and read $P(A \mid B)$ as "the probability of $A$ *given that* $B$
  happened," and explain why $P(A \mid B)$ and $P(B \mid A)$ are different quantities.
- Write Bayes' rule in discrete form and name each piece: prior, likelihood, marginal (evidence),
  posterior.
- Define **sensitivity**, **specificity**, and **base rate**, and combine them to get the
  probability of disease given a positive test.
- Solve a discrete updating problem **two ways** — as a counting table and via the Bayes' rule
  formula — and check that they agree.
- Diagnose **base-rate neglect**: spot when someone has confused $P(\text{positive} \mid \text{disease})$
  with $P(\text{disease} \mid \text{positive})$.

## Core vocabulary

- **Conditional probability** $P(A \mid B)$ — the probability of event $A$ restricted to the world
  where $B$ is true. Definition: $P(A \mid B) = P(A \cap B) / P(B)$ when $P(B) > 0$.
- **Prior** — what you believe about the unknown *before* seeing this piece of evidence. In the
  diagnostic case the prior is the **base rate**, $P(\text{disease})$.
- **Likelihood** — how probable the *evidence* is under each value of the unknown. For a test, this
  is sensitivity and the false-positive rate.
- **Marginal probability (evidence)** — the total probability of the evidence, summed over all values
  of the unknown. It is the normalizing piece $f(y)$ that makes the posterior a valid probability.
- **Posterior** — the updated belief about the unknown *after* the evidence, $P(\text{disease} \mid
  \text{positive})$.
- **Sensitivity** — $P(\text{positive} \mid \text{disease})$: among people who *have* the condition,
  the fraction the test correctly flags. (Also: true-positive rate.)
- **Specificity** — $P(\text{negative} \mid \text{no disease})$: among people who *do not* have it,
  the fraction the test correctly clears. (Its complement, $1 - \text{specificity}$, is the
  false-positive rate.)
- **Base rate** — the prior prevalence $P(\text{disease})$ in the population being tested.

## Conditional probability: the direction matters

The single most important idea this week is that **a conditional probability has a direction**, and
flipping the direction changes the answer. $P(\text{positive} \mid \text{disease})$ asks: *among sick
people, how many test positive?* That is a property of the test. $P(\text{disease} \mid \text{positive})$
asks: *among people who tested positive, how many are sick?* That is what a patient actually wants to
know. These two are related but not equal, and the gap between them is where Bayes' rule lives.

From the definition $P(A \mid B) = P(A \cap B)/P(B)$, notice that $P(A \cap B) = P(B \mid A)\,P(A)$.
Substituting gives the discrete form of **Bayes' rule**:

$$
P(A \mid B) \;=\; \frac{P(B \mid A)\, P(A)}{P(B)}.
$$

Read it as: posterior = (likelihood × prior) / evidence. The numerator reweights the prior $P(A)$ by
how well $A$ explains the evidence $B$; the denominator $P(B)$ rescales so the result is a genuine
probability. This is the discrete cousin of the course's central identity
$f(\theta \mid y) \propto L(\theta \mid y)\, f(\theta)$ — the marginal $P(B)$ (the $f(y)$) is exactly
the constant that $\propto$ would drop.

## The evidence term is a weighted sum

The denominator $P(B)$ rarely arrives handed to you. You build it by the **law of total probability**:
split the world into the possible values of the unknown, find the probability of the evidence under
each, and add. With a disease that is present ($D$) or absent ($D^c$):

$$
P(\text{positive}) \;=\; P(\text{positive} \mid D)\,P(D) \;+\; P(\text{positive} \mid D^c)\,P(D^c).
$$

The first term is **true positives** (sensitivity × base rate); the second is **false positives**
(false-positive rate × the much larger non-diseased fraction). When the disease is rare, that second
term can dominate — there are simply so many healthy people that even a small false-positive rate
produces a flood of positive tests from people who are fine. That is the whole mystery in one line.

## Sensitivity, specificity, base rate: putting them together

Plugging the test quantities into Bayes' rule:

$$
P(D \mid \text{positive}) \;=\;
\frac{\overbrace{P(\text{positive} \mid D)}^{\text{sensitivity}}\;\overbrace{P(D)}^{\text{base rate}}}
{\underbrace{P(\text{positive} \mid D)\,P(D)}_{\text{true positives}} +
\underbrace{P(\text{positive} \mid D^c)\,P(D^c)}_{\text{false positives}}}.
$$

Three knobs control the answer: a higher **base rate** raises the posterior, higher **sensitivity**
raises it, and higher **specificity** (fewer false positives) raises it. The base rate is the knob
people forget — and it is often the most powerful one when the condition is rare.

## Worked examples

The three cases below are the same machine run on different inputs: a rare-disease screen where the
posterior is surprisingly low, the recurring bike-to-campus survey as a discrete table, and a spam
filter where the posterior comes out high. Work each by hand and watch the base rate do its work.

### Worked example — synthetic disease screening (table *and* Bayes' rule)

Consider a **synthetic** screening scenario (numbers chosen for instruction, not from any real test
or dataset):

- Base rate: $P(D) = 0.01$ (1 in 100 has the condition).
- Sensitivity: $P(\text{positive} \mid D) = 0.95$.
- Specificity: $0.90$, so the false-positive rate is $P(\text{positive} \mid D^c) = 0.10$.

**As a counting table.** Imagine 10,000 people, which turns probabilities into whole-person counts
you can add up:

| | Test positive | Test negative | Row total |
|---|---|---|---|
| **Has condition** ($D$) | 95 | 5 | 100 |
| **No condition** ($D^c$) | 990 | 8910 | 9900 |
| **Column total** | 1085 | 8915 | 10000 |

Of the 100 diseased people, $0.95 \times 100 = 95$ test positive. Of the 9900 healthy people,
$0.10 \times 9900 = 990$ test positive. So among the **1085** total positives, only **95** truly have
the condition:

$$
P(D \mid \text{positive}) \;=\; \frac{95}{1085} \;\approx\; 0.0876.
$$

**Via Bayes' rule.** Same numbers, no table:

$$
P(D \mid \text{positive}) \;=\;
\frac{0.95 \times 0.01}{0.95 \times 0.01 + 0.10 \times 0.99}
\;=\; \frac{0.0095}{0.0095 + 0.099}
\;=\; \frac{0.0095}{0.1085} \;\approx\; 0.0876.
$$

The two methods agree, as they must — the table *is* the formula, written out person by person. The
headline: a "95% sensitive" test on a rare condition still leaves a positive person at under a **9%**
chance of actually having it. Most positives are false positives, because healthy people vastly
outnumber sick ones.

The figure below shows how the posterior $P(D \mid \text{positive})$ climbs as the base rate climbs,
holding sensitivity and specificity fixed at the values above.

```{r}
#| label: fig-baserate-curve
#| fig-cap: "Posterior probability of the condition given a positive test, as a function of the base rate (prevalence), with sensitivity fixed at 0.95 and false-positive rate at 0.10. The dot marks the worked example at a 1% base rate."
#| fig-alt: "A rising curve on axes where the horizontal axis is the base rate from 0 to 0.3 and the vertical axis is the posterior probability of disease given a positive test from 0 to 1. The curve starts near zero, rises steeply at first then more gently, passing through about 0.09 at a base rate of 0.01 where a marked dot sits, and reaching roughly 0.8 by a base rate of 0.3."
sens <- 0.95          # sensitivity P(positive | D)
fpr  <- 0.10          # false-positive rate P(positive | not D)
base <- seq(0.001, 0.30, length.out = 400)
post <- (sens * base) / (sens * base + fpr * (1 - base))
plot(base, post, type = "l", lwd = 2,
     xlab = "Base rate  P(D)",
     ylab = "Posterior  P(D | positive)",
     main = "A positive test means more when the condition is common",
     ylim = c(0, 1))
# mark the worked example at base rate 0.01
b0 <- 0.01
p0 <- (sens * b0) / (sens * b0 + fpr * (1 - b0))
points(b0, p0, pch = 19)
text(b0, p0, labels = sprintf("  base=0.01\n  post=%.3f", p0), pos = 4)
abline(h = 0.5, lty = 3)
legend("bottomright", legend = c("posterior curve", "0.5 reference"),
       lwd = c(2, 1), lty = c(1, 3), bty = "n")
```

Two lessons sit in this curve. The posterior is steeply sensitive to the base rate when the condition
is rare (the left side), and it only crosses the 0.5 line once the condition is fairly common. Same
test, very different meaning of a positive — entirely because of the prior.

### Worked example — recurring case: bike-to-campus, discrete updating table

Now the course's recurring question, in discrete form. We want the **proportion** $p$ of students who
bike to campus. Suppose, for this week only, that $p$ can take just a few candidate values, and we
hold a discrete **prior** over them (this is a coarse stand-in for the smooth Beta prior we adopt in
Week 3). Our evidence is a tiny pilot: we ask **3** students and **1** of them bikes.

For each candidate $p$, the likelihood of "1 biker out of 3" under independent responses is the
binomial weight $\binom{3}{1} p^1 (1-p)^2 = 3p(1-p)^2$. Because the $\binom{3}{1} = 3$ is the same for
every candidate, it cancels in Bayes' rule — a first taste of "$\propto$ drops constants." The
updating table:

| Candidate $p$ | Prior $f(p)$ | Likelihood $\propto p(1-p)^2$ | Prior × Likelihood | Posterior $f(p \mid y)$ |
|---|---|---|---|---|
| 0.2 | 0.20 | 0.128 | 0.02560 | 0.186 |
| 0.3 | 0.30 | 0.147 | 0.04410 | 0.320 |
| 0.4 | 0.30 | 0.144 | 0.04320 | 0.313 |
| 0.5 | 0.20 | 0.125 | 0.02500 | 0.181 |

The marginal (evidence) is the column sum of "Prior × Likelihood":
$0.02560 + 0.04410 + 0.04320 + 0.02500 = 0.13790$. Dividing each entry by 0.13790 gives the posterior
column above: $0.02560/0.13790 \approx 0.186$, $0.04410/0.13790 \approx 0.320$,
$0.04320/0.13790 \approx 0.313$, and $0.02500/0.13790 \approx 0.181$ — and as a check, those four
posterior weights sum to 1.

Notice the shape of the update: the prior favored $p = 0.3$ and $p = 0.4$ equally, and seeing 1
biker in 3 (an observed rate of $0.33$) concentrates a little more weight on the two candidates
nearest that rate — both $0.3$ and $0.4$ rise above their prior, while the extremes $0.2$ and $0.5$
fall below theirs. The data nudged our belief; it did not overturn it, because three observations
carry little weight. In Week 3
we let $p$ range over the whole interval $[0,1]$ and replace this table with a Beta prior — and the
*same* prior-times-likelihood logic produces the posterior **Beta(10, 18)** from the term's recurring
survey (prior Beta(2,2), 8 bikers of 24).

```{r}
#| label: fig-bike-discrete-update
#| fig-cap: "Discrete prior versus posterior over four candidate values of the bike-to-campus proportion, after observing 1 biker out of 3 surveyed. The posterior shifts mass slightly toward lower values."
#| fig-alt: "A grouped bar chart with four candidate proportions 0.2, 0.3, 0.4, 0.5 on the horizontal axis. For each candidate, a prior bar and a posterior bar stand side by side on a probability axis from 0 to about 0.35. The prior is tied highest at 0.3 and 0.4; the posterior is highest at 0.3 with 0.4 close behind, rising above the prior at 0.3 and 0.4 and falling below the prior at the extremes 0.2 and 0.5."
p_vals <- c(0.2, 0.3, 0.4, 0.5)
prior  <- c(0.20, 0.30, 0.30, 0.20)
lik    <- p_vals * (1 - p_vals)^2     # proportional to p(1-p)^2 (the choose(3,1) cancels)
unnorm <- prior * lik
post   <- unnorm / sum(unnorm)
mat <- rbind(prior, post)
barplot(mat, beside = TRUE, names.arg = p_vals,
        col = c("grey70", "grey30"),
        xlab = "Candidate proportion p (bike to campus)",
        ylab = "Probability weight",
        main = "Discrete updating: prior vs posterior",
        ylim = c(0, 0.55))
legend("topright", legend = c("prior", "posterior"),
       fill = c("grey70", "grey30"), bty = "n")
```

### Worked example — transfer case: spam filter (a positive flag)

The same arithmetic governs a context with nothing medical about it. A **synthetic** email filter
flags messages as "spam." Suppose 20% of incoming mail is genuinely spam ($P(S) = 0.20$, the base
rate). The filter flags 98% of true spam ($P(\text{flag} \mid S) = 0.98$) but also wrongly flags 3%
of legitimate mail ($P(\text{flag} \mid S^c) = 0.03$). A message lands in your spam folder — how
likely is it really spam?

$$
P(S \mid \text{flag}) \;=\;
\frac{0.98 \times 0.20}{0.98 \times 0.20 + 0.03 \times 0.80}
\;=\; \frac{0.196}{0.196 + 0.024}
\;=\; \frac{0.196}{0.220} \;\approx\; 0.891.
$$

Here the posterior is high — about **89%** — because the base rate (20%) is far from tiny and the
false-positive rate is low. Same formula, opposite feeling from the rare-disease case. The lesson
transfers exactly: a positive flag is only as trustworthy as the base rate and the false-positive
rate together allow.

## A common mistake

The classic error this week is **base-rate neglect**: treating the test's accuracy *as if* it were
the probability of disease. A patient hears "the test is 95% accurate, and mine was positive," and
concludes "so I'm 95% likely to be sick." That swaps $P(\text{positive} \mid D)$ — a property of the
test — for $P(D \mid \text{positive})$ — what they actually want. The two differ by exactly the
factor Bayes' rule supplies, and when the base rate is small they differ enormously (95% versus under
9% in our worked example).

**How to catch it.** Whenever you see a conditional probability, say out loud which event is the
"given." If the quoted number conditions on *disease status* but the question asks about *test
result* (or vice versa), you are about to flip a conditional and must run Bayes' rule. A reliable
gut-check: build the counting table for 10,000 people. If most of the positives come from the large
healthy group, the posterior is low no matter how "accurate" the test sounds.

## Interpretation guidance

A posterior like $P(D \mid \text{positive}) \approx 0.09$ does **not** mean the test is broken or
useless. It means a single positive on a rare condition is a *reason to investigate further*, not a
diagnosis — which is precisely why real screening programs confirm positives with a second,
independent test. (Re-running Bayes' rule with the first posterior as the new prior, and a second
positive as new evidence, drives the probability up sharply; that is Bayesian updating done twice.)

Equally, the number is **conditional on the model inputs**: the base rate, sensitivity, and
specificity we assumed. Change the population (and thus the base rate) and the posterior changes even
with the identical test. A posterior probability is always a statement *given* a prior and a
likelihood — never a free-standing fact about the world. Carry that habit forward: in this course we
will never report an estimate without being explicit about what it was conditioned on.

## Practice (ungraded)

Use these to check your understanding. No solutions are posted here.

1. A **synthetic** test has sensitivity 0.90 and specificity 0.95. The base rate is 0.02. Build the
   10,000-person counting table, then compute $P(D \mid \text{positive})$ with Bayes' rule, and
   confirm the two agree.
2. Holding sensitivity and specificity fixed, what base rate would push $P(D \mid \text{positive})$
   above 0.5? Reason from the curve in @fig-baserate-curve before computing.
3. In the bike-to-campus discrete table, finish the two missing posterior entries and verify the four
   posterior weights sum to 1. Which candidate gained the most weight, and why?
4. A friend says "the spam filter is 98% accurate, so anything in my spam folder is 98% likely to be
   spam." Name the misconception and explain, in one sentence, what number they actually computed.
5. Suppose you get a *second* independent positive on the rare-disease test. Use your Week-2 posterior
   as the new prior and update again. Roughly how much does the probability move?

## Reading guide

This week maps to **Bayes Rules! Chapter 2 (Bayes' Rule)**. Read it for the conditional-probability
groundwork and the diagnostic-test framing; our worked numbers and the bike-to-campus table are
course-original, so use the chapter for the *concepts* and this page for the *running examples*.

- The chapter's treatment of conditional probability and the difference between $P(A \mid B)$ and
  $P(B \mid A)$ underwrites our "**Conditional probability: the direction matters**" section — read
  it first.
- Its development of Bayes' rule (prior, likelihood, the normalizing/marginal term) maps onto "**The
  evidence term is a weighted sum**" and "**Sensitivity, specificity, base rate**" here. Pay
  attention to how the chapter builds the denominator by total probability — that is our weighted-sum
  point.
- Where the chapter discusses how a test result revises a probability, connect it to both worked
  diagnostic examples and to the recurring discrete updating table — the table is the chapter's logic
  rendered as counting.

## Public vs. graded

These notes and the ungraded practice above are the **public** face of the week. The graded quiz for
this week — its exact prompts, options, point values, rubric, and due date — lives in the LMS:
**the LMS (Blackboard) is authoritative** for everything graded. **No answer keys are posted here**,
and nothing on this page is a submission contract. Prepare from the worked examples and practice
prompts; submit and be graded in Blackboard.

## Looking ahead

Next week we let the unknown stop being a short list of candidates and become a *continuous* quantity.
**Week 3 — Prior, likelihood, posterior** replaces this week's discrete updating table with smooth
prior and posterior *densities*, and shows how the very same prior-times-likelihood move produces a
continuous posterior. The bike-to-campus table you built here becomes a Beta prior there.

## See also

- [Lab 4 — Beta-Binomial simulation](../labs/lab-04-beta-binomial-simulation.qmd) — the computational
  follow-through once the proportion goes continuous.
- [Notation glossary](../resources/notation-glossary.qmd) — the symbols and conventions this page
  inherits.
- [Bayes vs. classical cheat sheet](../resources/bayes-vs-classical-cheatsheet.qmd) — where credible
  intervals and confidence intervals part ways.
- [Week 3 — Prior, likelihood, posterior](../notes/week-03-priors-likelihoods-posteriors.qmd) — the
  continuous sequel.
- [Syllabus](../syllabus.qmd) — course shape and the public/graded boundary.
